
```{r}
getwd()

```


```{r}
#set working directory
setwd("C:/Users/yarram/Documents/Rna-seq-metadata-gap-filling/barley_rna_seq_analysis")

```




```{r}
#load necessary libraries
#loading libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(umap)
library(sva)
library(limma)
library(factoextra)
library(pheatmap)


```





```{r}
library(ggplot2)

# Create a summary data frame with the provided data
# here, i manually creating a summary dataframe from the results obtained from previous analysis
#so that we can use it for plotting to compare test and CV results of Random Forest model performance vs number of PCs

summary_df <- data.frame(
  PCs = c(10, 20, 50, 100),

  test_acc   = c(0.8363, 0.8272, 0.8573, 0.8573),
  test_kappa = c(0.7794, 0.7699, 0.8075, 0.8083),

  cv_acc     = c(0.7629103, 0.7629103, 0.7605146, 0.7613606),
  cv_acc_sd  = c(0.2001379, 0.2001379, 0.2114176, 0.2062091),

  cv_kappa    = c(0.6728584, 0.6728584, 0.6765530, 0.6787282),
  cv_kappa_sd = c(0.2489090, 0.2489090, 0.2534176, 0.2406590)
)

summary_df


```





```{r}
#line plots for accuracy and kappa vs number of PCs

p_acc <- ggplot(summary_df, aes(x = PCs)) +
  geom_point(aes(y = test_acc, shape = "Test accuracy"), size = 3) +
  geom_line(aes(y = test_acc, linetype = "Test accuracy")) +

  geom_point(aes(y = cv_acc, shape = "CV mean accuracy"), size = 3) +
  geom_line(aes(y = cv_acc, linetype = "CV mean accuracy")) +
  geom_errorbar(aes(ymin = cv_acc - cv_acc_sd, ymax = cv_acc + cv_acc_sd),
                width = 4, alpha = 0.7) +

  scale_x_continuous(breaks = summary_df$PCs) +
  labs(
    title = "Random Forest performance vs number of PCs (Tissue, SMART split)",
    x = "Number of PCs used",
    y = "Accuracy",
    shape = "",
    linetype = ""
  ) +
  theme_minimal()

p_acc



```

#Test accuracy increases slightly from PC10 → PC50

#After ~50 PCs, performance plateaus

#CV mean accuracy is:

#lower than test accuracy (this is GOOD and expected)

#very similar for PC20, PC50, PC100

#CV standard deviation is large → indicates heterogeneity across studies






```{r}
# Kappa plot
p_kappa <- ggplot(summary_df, aes(x = PCs)) +
  geom_point(aes(y = test_kappa, shape = "Test kappa"), size = 3) +
  geom_line(aes(y = test_kappa, linetype = "Test kappa")) +

  geom_point(aes(y = cv_kappa, shape = "CV mean kappa"), size = 3) +
  geom_line(aes(y = cv_kappa, linetype = "CV mean kappa")) +
  geom_errorbar(aes(ymin = cv_kappa - cv_kappa_sd, ymax = cv_kappa + cv_kappa_sd),
                width = 4, alpha = 0.7) +

  scale_x_continuous(breaks = summary_df$PCs) +
  labs(
    title = "Random Forest kappa vs number of PCs (Tissue, SMART split)",
    x = "Number of PCs used",
    y = "Kappa",
    shape = "",
    linetype = ""
  ) +
  theme_minimal()

p_kappa


```



#Kappa measures agreement beyond random guessing

#Rough interpretation:
#0.6–0.8 → substantial
#0.8 → almost perfect

#Test kappa increases from PC10 → PC50
#Plateaus at PC50–PC100
#CV kappa stays around ~0.67–0.69#Kappa measures agreement beyond random guessing

#Rough interpretation:
#0.6–0.8 → substantial
#0.8 → almost perfect

#Test kappa increases from PC10 → PC50
#Plateaus at PC50–PC100
#CV kappa stays around ~0.67–0.69




```{r}
# save plots to figures folder
ggsave(
  file.path("C:", "Users", "yarram", "Documents",
            "Rna-seq-metadata-gap-filling",
            "barley_rna_seq_analysis",
            "figures",
            "rf_performance_vs_pcs_tissue_smart_split_accuracy.png"),
  plot = p_acc, width = 8, height = 6
)



```




```{r}
# Create a summary data frame with the provided data
summary_df <- data.frame(
  PCs = c(10, 20, 50, 100),

  test_acc   = c(0.8363, 0.8272, 0.8573, 0.8573),
  test_kappa = c(0.7794, 0.7699, 0.8075, 0.8083),

  cv_acc     = c(0.7629103, 0.7629103, 0.7605146, 0.7613606),
  cv_acc_sd  = c(0.2001379, 0.2001379, 0.2114176, 0.2062091),

  cv_kappa    = c(0.6728584, 0.6728584, 0.6765530, 0.6787282),
  cv_kappa_sd = c(0.2489090, 0.2489090, 0.2534176, 0.2406590)
)

```





```{r}
library(ggplot2)

# Make sure PCs are treated in the correct order
summary_df$PCs <- factor(summary_df$PCs, levels = c(10, 20, 50, 100))

# Accuracy vs PCs plot

p_acc <- ggplot(summary_df, aes(x = PCs, group = 1)) +
  # CV mean accuracy + error bars (± SD)
  geom_errorbar(aes(ymin = cv_acc - cv_acc_sd, ymax = cv_acc + cv_acc_sd),
                width = 0.15, alpha = 0.6) +
  geom_line(aes(y = cv_acc, linetype = "CV mean"), linewidth = 0.7) +
  geom_point(aes(y = cv_acc, shape = "CV mean"), size = 3) +

  # Test accuracy
  geom_line(aes(y = test_acc, linetype = "Test"), linewidth = 0.7) +
  geom_point(aes(y = test_acc, shape = "Test"), size = 3) +

  # same y-scale across plots
  ylim(0, 1) +
  labs(
    title = "Model performance over PCs for Tissue (Accuracy)",
    x = "Number of PCs used",
    y = "Accuracy",
    shape = "",
    linetype = ""
  ) +
  theme_minimal()

p_acc

```




```{r}
# Kappa vs PCs plot

p_kappa <- ggplot(summary_df, aes(x = PCs, group = 1)) +
  # CV mean kappa + error bars (± SD)
  geom_errorbar(aes(ymin = cv_kappa - cv_kappa_sd, ymax = cv_kappa + cv_kappa_sd),
                width = 0.15, alpha = 0.6) +
  geom_line(aes(y = cv_kappa, linetype = "CV mean"), linewidth = 0.7) +
  geom_point(aes(y = cv_kappa, shape = "CV mean"), size = 3) +

  # Test kappa
  geom_line(aes(y = test_kappa, linetype = "Test"), linewidth = 0.7) +
  geom_point(aes(y = test_kappa, shape = "Test"), size = 3) +

  # same y-scale across plots
  ylim(0, 1) +
  labs(
    title = "Model performance over PCs for Tissue (Kappa)",
    x = "Number of PCs used",
    y = "Kappa",
    shape = "",
    linetype = ""
  ) +
  theme_minimal()

p_kappa


```


```{r}
## save plots to figures folder
ggsave(
  file.path("C:", "Users", "yarram", "Documents",
            "Rna-seq-metadata-gap-filling",
            "barley_rna_seq_analysis",
            "figures",
            "rf_performance_vs_pcs_tissue_smart_split_accuracy.png"),
  plot = p_acc, width = 8, height = 6
)
```


```{r}
```

