
```{r}

getwd()

```



```{r}
#set working directory
setwd("C:/Users/yarram/Documents/Rna-seq-metadata-gap-filling/barley_rna_seq_data")

```




```{r}
#load necessary libraries
#loading libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(umap)
library(sva)
library(limma)
library(factoextra)
library(pheatmap)

```





```{r}
# read in the four parts of the  data
df <- read.delim("../barley_rna_seq_data/my_barley_input.2.part1.abundance.tsv.gz", header = TRUE, sep = "\t")
df1 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part2.abundance.tsv.gz", header = TRUE, sep = "\t")
df2 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part3.abundance.tsv.gz", header = TRUE, sep = "\t")
df3 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part4.abundance.tsv.gz", header = TRUE, sep = "\t")


```






```{r}
# Rename the first column of each df to "gene"
names(df)[1]   <- "gene"
names(df1)[1]  <- "gene"
names(df2)[1]  <- "gene"
names( df3)[1] <- "gene"


```





```{r}

library(dplyr)
# Merge data frames by "gene" column using full_join
df4 <- list(df, df1, df2, df3) |>
  Reduce(function(x, y) full_join(x, y, by = "gene"), x = _)

# (optional) if any duplicate sample names exist, make them unique
colnames(df4) <- make.unique(colnames(df4))



```





```{r}

sum(is.na(df4)) 


```




```{r}
# extract TPM columns + gene column
tpm_df <- df4 %>%
  select(gene, ends_with("_tpm"))


```





```{r}
# quick check
dim(tpm_df)
head(tpm_df[, 1:6])


```




```{r}
#set row names as gene names and remove gene column

rownames(tpm_df) <- tpm_df$gene
tpm_df <- tpm_df[, -1]
#convert to matrix
tpm_matrix <- as.matrix(tpm_df)
mode(tpm_matrix) <- "numeric"


```





```{r}

#check dimensions
dim(tpm_matrix)

```


```{r}
#check first few samples
head(tpm_matrix[, 1:6])
head(tpm_matrix[, 1:6])


```




```{r}

#remove _tpm from column names so that they match metadata
colnames(tpm_matrix) <- gsub("_tpm$", "", colnames(tpm_matrix))


```





```{r}
#load your metadata
library(readxl)
meta_raw <- read_xlsx("hordeum_vulgare_10520.xlsx")

```





```{r}
# what are column names
colnames(meta_raw)


```



```{r}

# i want to see how many sheets of data are in the excel file
excel_sheets("hordeum_vulgare_10520.xlsx")


```




```{r}
#load necessary libraries before data cleaning
library(dplyr)
library(magrittr)  # makes sure %>% is available


```





```{r}
#i want to read in a specific sheet from the excel file
meta_raw <- read_xlsx("hordeum_vulgare_10520.xlsx", sheet = "Sheet1")


```





```{r}
#clean/standardize metadata (use your real columns)
meta <- meta_raw %>%
  rename(
    sample_id = Run,     # SRR/DRR/ERR
    tissue    = sample   # your tissue column is named sample
  ) %>%
  mutate(
    sample_id = as.character(sample_id),
    treatment = trimws(as.character(treatment)),
    layout    = tolower(trimws(as.character(layout))),
    tissue    = tolower(trimws(as.character(tissue)))
  ) %>%
  distinct(sample_id, .keep_all = TRUE)



```






```{r}
# checking the data
dim(meta)
head(meta[, c("sample_id","treatment","tissue","layout")])


```





```{r}
# keep only samples that are present in the TPM matrix, one row per sample_id
meta_keep <- meta %>%
  filter(sample_id %in% colnames(tpm_matrix)) %>%
  distinct(sample_id, .keep_all = TRUE)


```





```{r}
# order rows of metadata to match the TPM column order
ord  <- match(colnames(tpm_matrix), meta_keep$sample_id)
keep <- !is.na(ord)


```





```{r}
# apply the same filter/order to both objects
tpm_matrix   <- tpm_matrix[, keep, drop = FALSE]
meta_aligned <- meta_keep[ord[keep], ]


```





```{r}
# sanity checks
ncol(tpm_matrix)            # should equal:
nrow(meta_aligned)          # ...this
sum(is.na(meta_aligned$sample_id))  # should be 0
head(meta_aligned[, c("sample_id","treatment","tissue","layout")])


```




```{r}
# log1p-transform and transpose -> rows = samples, cols = genes
# RNA-seq TPM values can vary hugely — some genes have TPMs of 0.1, others 10,000+.
#Without transforming, the few highly expressed genes dominate the analysis.
#It stabilizes variance (so big and small expression values are comparable).

log_tpm <- t(log1p(tpm_matrix))   # big object: 10096 x ~37962


```





```{r}
# remove genes with zero variance (constant across all samples)
# This finds genes that have no change across samples — i.e., their expression is constant everywhere.
# Removing zero-variance genes keeps only the informative ones — genes that actually differ between samples — improving PCA/UMAP clarity and speed.
gene_var <- apply(log_tpm, 2, var)
sum_zero_var <- sum(gene_var == 0)         # how many genes removed?
log_tpm_f <- log_tpm[, gene_var > 0, drop = FALSE]


```






```{r}
#  quick checks
ncol_before <- ncol(log_tpm)
ncol_after  <- ncol(log_tpm_f)
cat("genes before:", ncol_before, "\ngenes after:", ncol_after,
    "\nzero-variance genes removed:", sum_zero_var, "\n")


```



```{r}
# run PCA on log-transformed TPM
pca_res <- prcomp(log_tpm_f, center = TRUE, scale. = TRUE)


```





```{r}
# variance explained by first few PCs
pc_var <- summary(pca_res)$importance[2, 1:6] * 100
pc_var


```




```{r}
library(dplyr)

# Clean metadata
# in this step, i created new columns for sra_study and tissue 

meta2 <- meta_aligned %>%
  mutate(
    sra_study = `SRA Study`,
    tissue = as.character(tissue)
  )



```





```{r}
# Keep only samples with known tissue + known study

keep_known <- !is.na(meta2$tissue) & meta2$tissue != "" &
              !is.na(meta2$sra_study) & meta2$sra_study != ""

meta_known <- meta2[keep_known, , drop = FALSE]


```





```{r}
# Features: first 20 PCs
# in this step, I extracted the first 20 principal components (PCs) from the PCA results for samples with known tissue labels.
# These PCs will be used as features for training a machine learning model to predict tissue types.
# y_all contains the corresponding tissue labels for these samples.

X20_all <- as.data.frame(pca_res$x[keep_known, 1:20, drop = FALSE])
y_all   <- factor(meta_known$tissue)


```





```{r}
# REMOVE "mixed" from training/evaluation dataset
# In this step, I filtered out samples labeled as "mixed" tissue from both the feature set (X20_all) and the labels (y_all).
# This ensures that the model is trained and evaluated only on samples with specific tissue types, improving classification accuracy.
# The resulting X20 and y variables contain only non-mixed tissue samples.


keep_nomixed <- y_all != "mixed"
X20 <- X20_all[keep_nomixed, , drop = FALSE]
y   <- droplevels(y_all[keep_nomixed])
meta_nomixed <- meta_known[keep_nomixed, , drop = FALSE]


```





```{r}
# Sanity checks
dim(X20)                       # samples x 20
table(y)                       # tissue counts (no mixed)
length(unique(meta_nomixed$sra_study))  # number of studies


```




```{r}
# Split data into training and test sets based on SRA studies
# In this step, I split the dataset into training and test sets based on unique SRA studies.
# This ensures that samples from the same study do not appear in both sets, preventing data leakage and providing a more realistic evaluation of model performance.
# Approximately 20% of the studies were randomly selected for the test set, while the remaining 80% were used for training.
# why am saying the 20% of studies? because samples from the same study may be more similar to each other than to samples from different studies.
# By splitting based on studies, we ensure that the model is tested on truly independent data, providing a better assessment of its generalization ability.
# This approach is particularly important in biological datasets where batch effects and study-specific biases can influence results.
#x_train, y_train: training features and labels
#x_test, y_test: test features and labels
# This split allows for robust model training and evaluation.

set.seed(123)

studies <- unique(meta_nomixed$sra_study)
test_studies <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_nomixed$sra_study %in% train_studies
test_idx  <- meta_nomixed$sra_study %in% test_studies

X_train <- X20[train_idx, , drop = FALSE]
y_train <- y[train_idx]

X_test  <- X20[test_idx, , drop = FALSE]
y_test  <- y[test_idx]

dim(X_train); dim(X_test)
length(unique(meta_nomixed$sra_study[train_idx]))
length(unique(meta_nomixed$sra_study[test_idx]))
```



# i splitted known samples for training and testing based on SRA study so that no overlap of studies in training and testing

#total known samples are 5365 from 183 studies
#Training: 4636 samples and 20 pcs were used as features
#Testing:   729 samples and 20 pcs were used as features
# Training set has 146 studies and test set has 37 studies  
# total studies are 183 which are unique studies in the metadata and same studies appear in the 4636 samples
# 183 studies are unique studies each study may have multiple samples




```{r}

# Train random forest on first 20 PCs
# In this step, I trained a random forest classifier using the first 20 principal components (PCs) as features to predict tissue types.
# The model was trained on the training set (X_train, y_train) with 500 trees.
# Random forests are robust to overfitting and can handle complex relationships in the data, making them suitable for this multi-class classification task.
# The trained model (rf20) can then be used to predict tissue types on unseen data (test set) and evaluate its performance. 
# This approach leverages(holds) the dimensionality reduction from PCA while utilizing the strengths of random forests for classification.
# This step is crucial for building a predictive model that can generalize well to new samples based on their gene expression profiles summarized by the PCs.
#importance = TRUE argument allows us to assess which features (PCs) are most influential in the classification decisions made by the model.


# Train random forest model

library(randomForest)
library(caret)

set.seed(123)

rf20 <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
)


# Evaluate on test set
# In this step, I evaluated the performance of the trained random forest model (rf20) on the independent test set (X_test, y_test).
# Predictions were made using the model, and a confusion matrix was generated to assess classification accuracy across different tissue types.
# The confusion matrix provides insights into how well the model performed, highlighting areas of strength and potential misclassifications.
# This evaluation is crucial for understanding the model's generalization ability to unseen data and identifying any tissue types that may require further attention or improvement in classification.
# The results from this evaluation inform future model refinement and validation efforts.
# pred_test contains the predicted tissue labels for the test set samples.
# cm20 contains the confusion matrix summarizing the model's performance.
#accuracy, sensitivity, specificity, and other metrics can be derived from cm20 to quantify model performance.
# This step is essential for validating the effectiveness of the random forest classifier in predicting tissue types based on the PCA-reduced features.
#kappa statistic from the confusion matrix provides a measure of agreement between predicted and true labels, accounting for chance agreement,means it helps to understand how well the model is performing beyond random guessing.



pred_test <- predict(rf20, X_test)
cm20 <- confusionMatrix(pred_test, y_test)
cm20



```



#kappa value is 0.7699 means the model is performing well beyond random guessing.
# the accuracy of the model is 0.8272 means the model correctly predicts the tissue type for approximately 82.72% of the samples in the test set.




```{r}

# Probabilities per tissue 
# In this step, I obtained the predicted probabilities for each tissue type for the samples in the test set using the trained random forest model (rf20).
# The predict function with type = "prob" returns a matrix where each row corresponds to a test sample and each column corresponds to a tissue type.
# These probabilities indicate the model's confidence in assigning each sample to each tissue class.
# The resulting prob_test20 object allows for a more nuanced interpretation of the model's predictions, as it provides insights into the likelihood of each tissue type rather than just the most likely class.

prob_test20 <- predict(rf20, X_test, type = "prob")
head(prob_test20)


```





```{r}
#10 fold cross validation on training set to estimate performance
# Cross-validation on training set to estimate performance
# In this step, I implemented a 10-fold cross-validation procedure on the training set to estimate the performance of the random forest model in a robust manner.
# The cross-validation was designed to ensure that samples from the same SRA study were kept together
# in either the training or validation fold, preventing data leakage and providing a more realistic assessment of model generalization.
# For each fold, the model was trained on 9/10 of the data and validated on the remaining 1/10, with accuracy and kappa statistics recorded for each iteration.
# This approach allows for a comprehensive evaluation of the model's performance across different subsets of the data, helping to identify potential overfitting and variability in predictions.
# The final results provide insights into the expected accuracy and reliability of the random forest classifier when applied to new, unseen samples.
#folds: list of indices for each fold based on studies means that samples from the same study are grouped together.


library(randomForest)
library(caret)

set.seed(123)

# Create folds based on SRA studies (SMART CV)
studies <- unique(meta_nomixed$sra_study)

# Create 10 folds of STUDIES (not samples)

folds <- createFolds(studies, k = 10, list = TRUE)

```





```{r}
# To store results
#acc_vec, kappa_vec: vectors to store accuracy and kappa statistics for each fold.
# These vectors will be populated during the cross-validation loop to summarize model performance across all folds.

acc_vec <- numeric(length(folds))
kappa_vec <- numeric(length(folds))


```





```{r}

#  Cross-validation loop.

for (i in seq_along(folds)) {

  test_studies <- studies[folds[[i]]]
  train_idx <- !(meta_nomixed$sra_study %in% test_studies)
  test_idx  <-  (meta_nomixed$sra_study %in% test_studies)

  X_train <- X20[train_idx, , drop = FALSE]
  y_train <- y[train_idx]

  X_test  <- X20[test_idx, , drop = FALSE]
  y_test  <- y[test_idx]

  rf_cv <- randomForest(
    x = X_train,
    y = y_train, 
    ntree = 500
  )

preds <- predict(rf_cv, X_test)
  cm <- confusionMatrix(preds, y_test)

  acc_vec[i]   <- cm$overall["Accuracy"]
  kappa_vec[i] <- cm$overall["Kappa"]
}


```





```{r}
# Cross-validation summary


mean(acc_vec)
sd(acc_vec)

mean(kappa_vec)
sd(kappa_vec)



```





```{r}
#

```






```{r}
```






```{r}
```






```{r}
```






```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```

