

```{r}
getwd()


```





```{r}
#set working directory
setwd("C:/Users/yarram/Documents/Rna-seq-metadata-gap-filling/barley_rna_seq_data")


```






```{r}
#load necessary libraries
#loading libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(umap)
library(sva)
library(limma)
library(factoextra)
library(pheatmap)


```






```{r}

# read in the four parts of the  data
df <- read.delim("../barley_rna_seq_data/my_barley_input.2.part1.abundance.tsv.gz", header = TRUE, sep = "\t")
df1 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part2.abundance.tsv.gz", header = TRUE, sep = "\t")
df2 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part3.abundance.tsv.gz", header = TRUE, sep = "\t")
df3 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part4.abundance.tsv.gz", header = TRUE, sep = "\t")




```






```{r}

# Rename the first column of each df to "gene"
names(df)[1]   <- "gene"
names(df1)[1]  <- "gene"
names(df2)[1]  <- "gene"
names( df3)[1] <- "gene"


```






```{r}

library(dplyr)
# Merge data frames by "gene" column using full_join
df4 <- list(df, df1, df2, df3) |>
  Reduce(function(x, y) full_join(x, y, by = "gene"), x = _)

# (optional) if any duplicate sample names exist, make them unique
colnames(df4) <- make.unique(colnames(df4))



```







```{r}

sum(is.na(df4)) 

```







```{r}
# extract TPM columns + gene column
tpm_df <- df4 %>%
  select(gene, ends_with("_tpm")) 


```







```{r}

# quick check
dim(tpm_df)
head(tpm_df[, 1:6])


```








```{r}
#set row names as gene names and remove gene column

rownames(tpm_df) <- tpm_df$gene
tpm_df <- tpm_df[, -1]
#convert to matrix
tpm_matrix <- as.matrix(tpm_df)
mode(tpm_matrix) <- "numeric"





```







```{r}

#check dimensions
dim(tpm_matrix)



```







```{r}

#check first few samples
head(tpm_matrix[, 1:6])
head(tpm_matrix[, 1:6])


```







```{r}

#remove _tpm from column names so that they match metadata
colnames(tpm_matrix) <- gsub("_tpm$", "", colnames(tpm_matrix))


```







```{r}

#load your metadata
library(readxl)
meta_raw <- read_xlsx("hordeum_vulgare_10520.xlsx")


```







```{r}
# what are column names
colnames(meta_raw)


```







```{r}

# i want to see how many sheets of data are in the excel file
excel_sheets("hordeum_vulgare_10520.xlsx")


```







```{r}

#load necessary libraries before data cleaning
library(dplyr)
library(magrittr)  # makes sure %>% is available


```







```{r}
#i want to read in a specific sheet from the excel file
meta_raw <- read_xlsx("hordeum_vulgare_10520.xlsx", sheet = "Sheet1")


```






```{r}
#clean/standardize metadata (use your real columns)
meta <- meta_raw %>%
  rename(
    sample_id = Run,     # SRR/DRR/ERR
    tissue    = sample   # your tissue column is named sample
  ) %>%
  mutate(
    sample_id = as.character(sample_id),
    treatment = trimws(as.character(treatment)),
    layout    = tolower(trimws(as.character(layout))),
    tissue    = tolower(trimws(as.character(tissue)))
  ) %>%
  distinct(sample_id, .keep_all = TRUE)



```







```{r}
# checking the data
dim(meta)
head(meta[, c("sample_id","treatment","tissue","layout")])


```







```{r}
# keep only samples that are present in the TPM matrix, one row per sample_id
meta_keep <- meta %>%
  filter(sample_id %in% colnames(tpm_matrix)) %>%
  distinct(sample_id, .keep_all = TRUE)


```







```{r}

# order rows of metadata to match the TPM column order
ord  <- match(colnames(tpm_matrix), meta_keep$sample_id)
keep <- !is.na(ord)


```







```{r}

# apply the same filter/order to both objects
tpm_matrix   <- tpm_matrix[, keep, drop = FALSE]
meta_aligned <- meta_keep[ord[keep], ]


```







```{r}

# sanity checks
ncol(tpm_matrix)            # should equal:
nrow(meta_aligned)          # ...this
sum(is.na(meta_aligned$sample_id))  # should be 0
head(meta_aligned[, c("sample_id","treatment","tissue","layout")])



```






```{r}

# log1p-transform and transpose -> rows = samples, cols = genes
# RNA-seq TPM values can vary hugely — some genes have TPMs of 0.1, others 10,000+.
#Without transforming, the few highly expressed genes dominate the analysis.
#It stabilizes variance (so big and small expression values are comparable).

log_tpm <- t(log1p(tpm_matrix))   # big object: 10096 x ~37962


```







```{r}
# remove genes with zero variance (constant across all samples)
# This finds genes that have no change across samples — i.e., their expression is constant everywhere.
# Removing zero-variance genes keeps only the informative ones — genes that actually differ between samples — improving PCA/UMAP clarity and speed.
gene_var <- apply(log_tpm, 2, var)
sum_zero_var <- sum(gene_var == 0)         # how many genes removed?
log_tpm_f <- log_tpm[, gene_var > 0, drop = FALSE]



```








```{r}
#  quick checks
ncol_before <- ncol(log_tpm)
ncol_after  <- ncol(log_tpm_f)
cat("genes before:", ncol_before, "\ngenes after:", ncol_after,
    "\nzero-variance genes removed:", sum_zero_var, "\n")


```








```{r}

# run PCA on log-transformed TPM
pca_res <- prcomp(log_tpm_f, center = TRUE, scale. = TRUE)



```







```{r}

# variance explained by first few PCs
pc_var <- summary(pca_res)$importance[2, 1:6] * 100
pc_var



```

 





```{r}
library(dplyr)

# Clean metadata 

meta2 <- meta_aligned %>%
  mutate(
    sra_study = `SRA Study`,
    tissue = as.character(tissue)
  )


```







```{r}

# Keep only samples with known tissue + known study
keep_known <- !is.na(meta2$tissue) & meta2$tissue != "" &
              !is.na(meta2$sra_study) & meta2$sra_study != ""

meta_known <- meta2[keep_known, , drop = FALSE]



```







```{r}
# Extract PC1–PC50
X50_all <- as.data.frame(pca_res$x[keep_known, 1:50, drop = FALSE])
y_all   <- factor(meta_known$tissue)



```






```{r}

# Remove "mixed" from training
keep_nomixed <- y_all != "mixed"
X50 <- X50_all[keep_nomixed, , drop = FALSE]
y   <- droplevels(y_all[keep_nomixed])
meta_nomixed <- meta_known[keep_nomixed, , drop = FALSE]



```







```{r}

# Sanity checks
dim(X50)                       # samples x 50
table(y)                       # tissue counts (no mixed)
length(unique(meta_nomixed$sra_study))  # number of studies



```






```{r}
# Split into train/test by study
set.seed(123)

studies <- unique(meta_nomixed$sra_study)
test_studies <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_nomixed$sra_study %in% train_studies
test_idx  <- meta_nomixed$sra_study %in% test_studies

X_train <- X50[train_idx, , drop = FALSE]
y_train <- y[train_idx]

X_test  <- X50[test_idx, , drop = FALSE]
y_test  <- y[test_idx]

# Checks
dim(X_train); dim(X_test)
length(unique(meta_nomixed$sra_study[train_idx]))
length(unique(meta_nomixed$sra_study[test_idx]))


```



# i splitted known samples for training and testing based on SRA study so that no overlap of studies in training and testing

#total known samples are 5365 from 183 studies
#Training: 4636 samples and 20 pcs were used as features
#Testing:   729 samples and 20 pcs were used as features
# Training set has 146 studies and test set has 37 studies  
# total studies are 183 which are unique studies in the metadata and same studies appear in the 4636 samples
# 183 studies are unique studies each study may have multiple samples







```{r}

#
library(randomForest)
library(caret)

set.seed(123)

rf50 <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
)

pred_test <- predict(rf50, X_test)
cm50 <- confusionMatrix(pred_test, y_test)
cm50


```



#kappa results are 0.8075 means "substantial agreement" between predicted and actual tissue labels in the test set.meaning the model is performing well in classifying tissue types based on the first 50 principal components of the gene expression data.

#accuracy results are 0.8573 means that approximately 85.73% of the tissue type predictions made by the random forest model on the test set were correct.






```{r}

# Probabilities
prob_test50 <- predict(rf50, X_test, type = "prob")
head(prob_test50)


```







```{r}

library(randomForest)
library(caret)

set.seed(123)

# 1) Build folds on STUDIES (not samples)
studies <- unique(meta_nomixed$sra_study)
folds <- createFolds(studies, k = 10, list = TRUE)


```







```{r}
# Store fold metrics
acc_vec <- numeric(length(folds))
kappa_vec <- numeric(length(folds))


```







```{r}

# CV loop
for (i in seq_along(folds)) {

  test_studies <- studies[folds[[i]]]

  train_idx <- !(meta_nomixed$sra_study %in% test_studies)
  test_idx  <-  (meta_nomixed$sra_study %in% test_studies)

  X_train <- X50[train_idx, , drop = FALSE]
  y_train <- y[train_idx]

  X_test  <- X50[test_idx, , drop = FALSE]
  y_test  <- y[test_idx]

  rf_cv <- randomForest(
    x = X_train,
    y = y_train,
    ntree = 500
  )

  preds <- predict(rf_cv, X_test)
  cm <- confusionMatrix(preds, y_test)

  acc_vec[i]   <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[i] <- as.numeric(cm$overall["Kappa"])
}


```







```{r}
# Summarize CV results
mean(acc_vec)
sd(acc_vec)

mean(kappa_vec)
sd(kappa_vec)

```







```{r}




```







```{r}




```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```

