

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(umap)
library(sva)
library(limma)
library(factoextra)
library(pheatmap)


```


```{r}

getwd()

```


```{r}
#set working directory
setwd("C:/Users/yarram/Documents/Rna-seq-metadata-gap-filling/barley_rna_seq_data")

```




```{r}
# read in the four parts of the  data
df <- read.delim("../barley_rna_seq_data/my_barley_input.2.part1.abundance.tsv.gz", header = TRUE, sep = "\t")
df1 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part2.abundance.tsv.gz", header = TRUE, sep = "\t")
df2 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part3.abundance.tsv.gz", header = TRUE, sep = "\t")
df3 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part4.abundance.tsv.gz", header = TRUE, sep = "\t")

```





```{r}
# Rename the first column of each df to "gene"
names(df)[1]   <- "gene"
names(df1)[1]  <- "gene"
names(df2)[1]  <- "gene"
names( df3)[1] <- "gene"

```





```{r}

library(dplyr)
# Merge data frames by "gene" column using full_join
df4 <- list(df, df1, df2, df3) |>
  Reduce(function(x, y) full_join(x, y, by = "gene"), x = _)

# (optional) if any duplicate sample names exist, make them unique
colnames(df4) <- make.unique(colnames(df4))


```






```{r}
# check for missing values
sum(is.na(df4)) 


```



```{r}
# extract TPM columns + gene column
tpm_df <- df4 %>%
  select(gene, ends_with("_tpm")) 

```






```{r}
# quick check
dim(tpm_df)
head(tpm_df[, 1:6])


```





```{r}
#set row names as gene names and remove gene column

rownames(tpm_df) <- tpm_df$gene
tpm_df <- tpm_df[, -1]
#convert to matrix
tpm_matrix <- as.matrix(tpm_df)
mode(tpm_matrix) <- "numeric"

```





```{r}
#check dimensions
dim(tpm_matrix)

```




```{r}
#check first few samples
head(tpm_matrix[, 1:6])
head(tpm_matrix[, 1:6])

```




```{r}
# # check if specific genes of interest are present in the TPM matrix
# genes <- c(
#   "HORVU.MOREX.r3.4HG0392430.1",
#   "HORVU.MOREX.r3.4HG0392420.1",
#   "HORVU.MOREX.r3.4HG0390910.2",
#   "HORVU.MOREX.r3.4HG0390560.1",
#   "HORVU.MOREX.r3.4HG0309800.1"
# )
# 
# genes %in% rownames(tpm_matrix)

```




```{r}
# # genes and TPM values for those genes (if present)
# genes_present <- genes[genes %in% rownames(tpm_matrix)]
# tpm_matrix[genes_present, 1:6]
# 
# #which gene is missing?
# genes[!genes %in% rownames(tpm_matrix)]


```


```{r}
# genes_present <- genes[genes %in% rownames(tpm_matrix)]
# gene_tpm_table <- tpm_matrix[genes_present, ]
# 
# # optional: write table
# write.csv(gene_tpm_table, "selected_genes_TPM.csv")

```





```{r}
# log_tpm <- log2(gene_tpm_table + 1)

```




```{r}
# pheatmap::pheatmap(
#   log_tpm,
#   scale = "row",
#   clustering_distance_rows = "euclidean",
#   clustering_distance_cols = "euclidean",
#   clustering_method = "complete",
#   fontsize_row = 10,
#   fontsize_col = 8
# )

```




```{r}
# write.csv(
#   meta_aligned[, c("sample_id", "treatment", "tissue", "layout")],
#   "sample_metadata.csv",
#   row.names = FALSE
# )

```




```{r}
#remove _tpm from column names so that they match metadata
colnames(tpm_matrix) <- gsub("_tpm$", "", colnames(tpm_matrix))

```





```{r}
#load your metadata
library(readxl)
meta_raw <- read_xlsx("hordeum_vulgare_10520.xlsx")

```






```{r}
# what are column names
colnames(meta_raw)

```




```{r}
# i want to see how many sheets of data are in the excel file
excel_sheets("hordeum_vulgare_10520.xlsx")

```




```{r}
#load necessary libraries before data cleaning
library(dplyr)
library(magrittr)  # makes sure %>% is available


```






```{r}
#i want to read in a specific sheet from the excel file
meta_raw <- read_xlsx("hordeum_vulgare_10520.xlsx", sheet = "Sheet1")

```





```{r}

#clean/standardize metadata (use your real columns)
meta <- meta_raw %>%
  rename(
    sample_id = Run,     # SRR/DRR/ERR
    tissue    = sample   # your tissue column is named sample
  ) %>%
  mutate(
    sample_id = as.character(sample_id),
    treatment = trimws(as.character(treatment)),
    layout    = tolower(trimws(as.character(layout))),
    tissue    = tolower(trimws(as.character(tissue)))
  ) %>%
  distinct(sample_id, .keep_all = TRUE)


```






```{r}
# checking the data
dim(meta)
head(meta[, c("sample_id","treatment","tissue","layout")])


```






```{r}
# keep only samples that are present in the TPM matrix, one row per sample_id
meta_keep <- meta %>%
  filter(sample_id %in% colnames(tpm_matrix)) %>%
  distinct(sample_id, .keep_all = TRUE)


```






```{r}
# keep only samples that are present in the TPM matrix, one row per sample_id
meta_keep <- meta %>%
  filter(sample_id %in% colnames(tpm_matrix)) %>%
  distinct(sample_id, .keep_all = TRUE)

```






```{r}
# order rows of metadata to match the TPM column order
ord  <- match(colnames(tpm_matrix), meta_keep$sample_id)
keep <- !is.na(ord)


```






```{r}
# apply the same filter/order to both objects
tpm_matrix   <- tpm_matrix[, keep, drop = FALSE]
meta_aligned <- meta_keep[ord[keep], ]


```





```{r}
#save  the aligned metadata in excel format for future use
#load library for writing excel files

#library(openxlsx)
#write.xlsx(meta_aligned, "aligned_metadata.xlsx", rowNames = FALSE)

```





```{r}
# sanity checks
ncol(tpm_matrix)            # should equal:
nrow(meta_aligned)          # ...this
sum(is.na(meta_aligned$sample_id))  # should be 0
head(meta_aligned[, c("sample_id","treatment","tissue","layout")])

```






```{r}
# log1p-transform and transpose -> rows = samples, cols = genes
# RNA-seq TPM values can vary hugely — some genes have TPMs of 0.1, others 10,000+.
#Without transforming, the few highly expressed genes dominate the analysis.
#It stabilizes variance (so big and small expression values are comparable).

log_tpm <- t(log1p(tpm_matrix))   # big object: 10096 x ~37962


```






```{r}
# remove genes with zero variance (constant across all samples)
# This finds genes that have no change across samples — i.e., their expression is constant everywhere.
# Removing zero-variance genes keeps only the informative ones — genes that actually differ between samples — improving PCA/UMAP clarity and speed.
gene_var <- apply(log_tpm, 2, var)
sum_zero_var <- sum(gene_var == 0)         # how many genes removed?
log_tpm_f <- log_tpm[, gene_var > 0, drop = FALSE]


```






```{r}
#  quick checks
ncol_before <- ncol(log_tpm)
ncol_after  <- ncol(log_tpm_f)
cat("genes before:", ncol_before, "\ngenes after:", ncol_after,
    "\nzero-variance genes removed:", sum_zero_var, "\n")


```






```{r}
# run PCA on log-transformed TPM
pca_res <- prcomp(log_tpm_f, center = TRUE, scale. = TRUE)


```






```{r}
# variance explained by first few PCs
pc_var <- summary(pca_res)$importance[2, 1:6] * 100
pc_var


```





```{r}

library(dplyr)
meta2 <- meta_aligned %>%
  mutate(
    sra_study = `SRA Study`,
    developmental_stage = as.character(developmental_stage)
  )

```






```{r}

keep_known <- !is.na(meta2$developmental_stage) & meta2$developmental_stage != "" &
              !is.na(meta2$sra_study) & meta2$sra_study != ""

meta_known <- meta2[keep_known, , drop = FALSE]

```






```{r}
# keep only the first 50 PCs for samples with known developmental stage
X50 <- as.data.frame(pca_res$x[keep_known, 1:50, drop = FALSE])
y <- factor(meta_known$developmental_stage)

```





```{r}
#sanity check
dim(X50)  # should be (number of known samples) x 50
table(y) # should equal number of rows in X50_all
#number of unique studies in known samples
length(unique(meta_known$sra_study))

```




```{r}
library(randomForest)
library(caret)

set.seed(123)

# remove transition BEFORE modeling 
keep_stage <- y != "transition"
X50_stage <- X50[keep_stage, , drop = FALSE]
y_stage <- droplevels(y[keep_stage])
meta_stage <- meta_known[keep_stage, , drop = FALSE]

# study-aware 80/20 split
studies <- unique(meta_stage$sra_study)
test_studies <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_stage$sra_study %in% train_studies
test_idx  <- meta_stage$sra_study %in% test_studies

X_train <- X50_stage[train_idx, , drop = FALSE]
y_train <- y_stage[train_idx]

X_test  <- X50_stage[test_idx, , drop = FALSE]
y_test  <- y_stage[test_idx]

# sanity checks
dim(X_train); dim(X_test)
length(unique(meta_stage$sra_study[train_idx]))
length(unique(meta_stage$sra_study[test_idx]))


```




```{r}
 # train RF without transition by 50 PCs
rf_stage <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
)

# predictions + confusion matrix
pred_test <- predict(rf_stage, X_test)
cm_stage <- confusionMatrix(pred_test, y_test)
cm_stage



```





```{r}
# Now do the same thing but with all known samples (including transition) to predict developmental stage (including transition)

# set.seed(123)
# 
# studies <- unique(meta_known$sra_study)
# test_studies <- sample(studies, size = round(0.2 * length(studies)))
# train_studies <- setdiff(studies, test_studies)
# 
# train_idx <- meta_known$sra_study %in% train_studies
# test_idx  <- meta_known$sra_study %in% test_studies
# 
# X_train <- X50[train_idx, , drop = FALSE]
# y_train <- y[train_idx]
# 
# X_test  <- X50[test_idx, , drop = FALSE]
# y_test  <- y[test_idx]
# 
# dim(X_train); dim(X_test)
# length(unique(meta_known$sra_study[train_idx]))
# length(unique(meta_known$sra_study[test_idx]))


```




```{r}
# Train a random forest classifier using the first 50 PCs to predict treatment
# library(randomForest)
# library(caret)
# 
# set.seed(123)
# 
# rf50 <- randomForest(
#   x = X_train,
#   y = y_train,
#   ntree = 500,
#   importance = TRUE
# )
# 
# pred_test <- predict(rf50, X_test)
# cm50 <- confusionMatrix(pred_test, y_test)
# cm50


```
#including transition





```{r}
# #this is with transition included
# prob_test50 <- predict(rf50, X_test, type = "prob")
# head(prob_test50)


```
#including transition




```{r}
# Train a random forest classifier using the first 50 PCs to predict developmental stage (excluding transition)  
# prob_stage <- predict(rf_stage, X_test, type = "prob") makes the model output probabilities for each class instead of just the predicted class.

prob_stage <- predict(rf_stage, X_test, type = "prob")
head(prob_stage)

```
#excluding transition



```{r}
# 10-fold study-aware CV for developmental stage prediction (excluding transition)
set.seed(123)

studies <- unique(meta_stage$sra_study)
folds <- createFolds(studies, k = 10, list = TRUE)

acc_vec <- numeric(10)
kappa_vec <- numeric(10)

for (k in 1:10) {

  test_studies <- studies[folds[[k]]]

  test_idx  <- meta_stage$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X50_stage[train_idx, , drop = FALSE]
  y_train <- y_stage[train_idx]

  X_test  <- X50_stage[test_idx, , drop = FALSE]
  y_test  <- y_stage[test_idx]

  rf <- randomForest(
    x = X_train,
    y = y_train,
    ntree = 500
  )

  pred <- predict(rf, X_test)
  cm <- confusionMatrix(pred, y_test)

  acc_vec[k] <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

# CV summary
mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)


```





```{r}
# CV loop
# 
# for (k in 1:50) {
# 
#   test_studies <- study_folds$sra_study[study_folds$fold == k]
#   
#   test_idx  <- meta_known$sra_study %in% test_studies
#   train_idx <- !test_idx
#   
#   X_train <- X50[train_idx, ]
#   y_train <- y[train_idx]
#   
#   X_test <- X50[test_idx, ]
#   y_test <- y[test_idx]
#   
#   rf <- randomForest(
#     x = X_train,
#     y = y_train,
#     ntree = 500
#   )
#   
#   pred <- predict(rf, X_test)
#   
#   cm <- confusionMatrix(pred, y_test)
#   
#   acc_vec[k] <- cm$overall["Accuracy"]
#   kappa_vec[k] <- cm$overall["Kappa"]
# }


```



#10 PCs for developmental stage prediction (excluding transition)

```{r}
# keep only the first 10 PCs for samples with known developmental stage
X10 <- as.data.frame(pca_res$x[keep_known, 1:10, drop = FALSE])
y <- factor(meta_known$developmental_stage)

```




```{r}
library(randomForest)
library(caret)

set.seed(123)

# remove transition BEFORE modeling 
keep_stage <- y != "transition"
X10_stage <- X10[keep_stage, , drop = FALSE]
y_stage <- droplevels(y[keep_stage])
meta_stage <- meta_known[keep_stage, , drop = FALSE]

# study-aware 80/20 split
studies <- unique(meta_stage$sra_study)
test_studies <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_stage$sra_study %in% train_studies
test_idx  <- meta_stage$sra_study %in% test_studies

X_train <- X10_stage[train_idx, , drop = FALSE]
y_train <- y_stage[train_idx]

X_test  <- X10_stage[test_idx, , drop = FALSE]
y_test  <- y_stage[test_idx]

# sanity checks
dim(X_train); dim(X_test)
length(unique(meta_stage$sra_study[train_idx]))
length(unique(meta_stage$sra_study[test_idx]))



```




```{r}
 # train RF without transition by 10 PCs
rf_stage <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
)

# predictions + confusion matrix
pred_test <- predict(rf_stage, X_test)
cm_stage <- confusionMatrix(pred_test, y_test)
cm_stage


```




```{r}
#probabilities for the 10 PC model
prob_stage_10 <- predict(rf_stage, X_test, type = "prob")
head(prob_stage_10)

```


```{r}
# 10-fold CV with study-aware folds (excluding transition) for 10 PCs
set.seed(123)
studies <- unique(meta_stage$sra_study)
folds <- createFolds(studies, k = 10, list = TRUE)
acc_vec <- numeric(length(folds))
kappa_vec <- numeric(length(folds))

for ( k in 1:10) {
  test_studies <- studies[folds[[k]]]
  test_idx  <- meta_stage$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X10_stage[train_idx, , drop = FALSE]
  y_train <- y_stage[train_idx]

  X_test  <- X10_stage[test_idx, , drop = FALSE]
  y_test  <- y_stage[test_idx]

  rf_cv <- randomForest(
    x = X_train,
    y = y_train,
    ntree = 500
  )

  pred <- predict(rf_cv, X_test, type = "response")
  cm <- confusionMatrix(pred, y_test)

  acc_vec[k] <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)  

```





#20 PCs for developmental stage prediction (excluding transition)
```{r}
# keep only the first 20 PCs for samples with known developmental stage
X20 <- as.data.frame(pca_res$x[keep_known, 1:20, drop = FALSE])
y <- factor(meta_known$developmental_stage)

```





```{r}
library(randomForest)
library(caret)

set.seed(123)

# remove transition BEFORE modeling
keep_stage <- y != "transition"
X20_stage <- X20[keep_stage, , drop = FALSE]
y_stage <- droplevels(y[keep_stage])
meta_stage <- meta_known[keep_stage, , drop = FALSE]  

# study-aware 80/20 split
studies <- unique(meta_stage$sra_study)
test_studies <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_stage$sra_study %in% train_studies
test_idx  <- meta_stage$sra_study %in% test_studies

X_train <- X20_stage[train_idx, , drop = FALSE]
y_train <- y_stage[train_idx]

X_test  <- X20_stage[test_idx, , drop = FALSE]
y_test  <- y_stage[test_idx]  

# sanity checks
dim(X_train); dim(X_test)
length(unique(meta_stage$sra_study[train_idx]))
length(unique(meta_stage$sra_study[test_idx]))


```





```{r}
# train RF without transition by 20 PCs
rf_stage <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
)

# predictions + confusion matrix
pred_test <- predict(rf_stage, X_test)
cm_stage <- confusionMatrix(pred_test, y_test)
cm_stage


```







```{r}
# probabilities for the 20 PC model
prob_stage_20 <- predict(rf_stage, X_test, type = "prob")
head(prob_stage_20)

```





```{r}
# 10-fold CV with study-aware folds (excluding transition) for 20 PCs
set.seed(123)
studies <- unique(meta_stage$sra_study)
folds <- createFolds(studies, k = 10, list = TRUE)
acc_vec <- numeric(length(folds))
kappa_vec <- numeric(length(folds)) 

for (k in 1:10) {
  test_studies <- studies[folds[[k]]]
  test_idx  <- meta_stage$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X20_stage[train_idx, , drop = FALSE]
  y_train <- y_stage[train_idx]

  X_test  <- X20_stage[test_idx, , drop = FALSE]
  y_test  <- y_stage[test_idx]

  rf_cv <- randomForest(
    x = X_train,
    y = y_train,
    ntree = 500
  )

  pred <- predict(rf_cv, X_test, type = "response")
  cm <- confusionMatrix(pred, y_test)

  acc_vec[k] <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)  



```



#100 PCs for developmental stage prediction (excluding transition)  

```{r}
# keep only the first 100 PCs for samples with known developmental stage
X100 <- as.data.frame(pca_res$x[keep_known, 1:100, drop = FALSE])
y <- factor(meta_known$developmental_stage)

```





```{r}
library(randomForest)
library(caret)

set.seed(123) 
  
# remove transition BEFORE modeling
keep_stage <- y != "transition"
X100_stage <- X100[keep_stage, , drop = FALSE]
y_stage <- droplevels(y[keep_stage])
meta_stage <- meta_known[keep_stage, , drop = FALSE]

# study-aware 80/20 split
studies <- unique(meta_stage$sra_study)
test_studies <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies) 

train_idx <- meta_stage$sra_study %in% train_studies
test_idx  <- meta_stage$sra_study %in% test_studies

X_train <- X100_stage[train_idx, , drop = FALSE]
y_train <- y_stage[train_idx]

X_test  <- X100_stage[test_idx, , drop = FALSE]
y_test  <- y_stage[test_idx]

# sanity checks
dim(X_train); dim(X_test)
length(unique(meta_stage$sra_study[train_idx]))
length(unique(meta_stage$sra_study[test_idx]))


```





```{r}
# train RF without transition by 100 PCs
rf_stage <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
) 

# predictions + confusion matrix
pred_test <- predict(rf_stage, X_test)
cm_stage <- confusionMatrix(pred_test, y_test)
cm_stage

```






```{r}
# probabilities for the 100 PC model
prob_stage_100 <- predict(rf_stage, X_test, type = "prob")
head(prob_stage_100)

```





```{r}
# 10-fold CV with study-aware folds (excluding transition) for 100 PCs
set.seed(123)
studies <- unique(meta_stage$sra_study)
folds <- createFolds(studies, k = 10, list = TRUE)
acc_vec <- numeric(length(folds))
kappa_vec <- numeric(length(folds)) 

for (k in 1:10) {
  test_studies <- studies[folds[[k]]]
  test_idx  <- meta_stage$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X100_stage[train_idx, , drop = FALSE]
  y_train <- y_stage[train_idx]

  X_test  <- X100_stage[test_idx, , drop = FALSE]
  y_test  <- y_stage[test_idx]

  rf_cv <- randomForest(
    x = X_train,
    y = y_train,
    ntree = 500
  )

  pred <- predict(rf_cv, X_test, type = "response")
  cm <- confusionMatrix(pred, y_test)

  acc_vec[k] <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)  


```




# NAIVE way by 50 PCs for developmental stage prediction (excluding transition) - random 80/20 split by sample, not by study 

```{r}
# Stupid split: random 80/20 by sample (not by study)
library(randomForest)
library(caret)

set.seed(123)

n <- nrow(X50_stage)
train_idx <- sample(seq_len(n), size = round(0.8 * n))
test_idx  <- setdiff(seq_len(n), train_idx)

X_train <- X50_stage[train_idx, , drop = FALSE]
y_train <- y_stage[train_idx]

X_test  <- X50_stage[test_idx, , drop = FALSE]
y_test  <- y_stage[test_idx]

dim(X_train); dim(X_test)



```




```{r}
# sanity checks
length(y_stage)

```

#2986 + 747 = 3733 total samples with known developmental stage excluding transition. So 80% of 3733 is about 2986 samples in training and 747 in testing.




```{r}
# random forest with stupid split 
set.seed(123)

rf50_stage_stupid <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
)

pred_test <- predict(rf50_stage_stupid, X_test)
cm_stage_stupid <- confusionMatrix(pred_test, y_test)
cm_stage_stupid


```






```{r}
# probabilities for the stupid split model
prob_stage_stupid <- predict(rf50_stage_stupid, X_test, type = "prob")
head(prob_stage_stupid)


```





```{r}
# 10-fold CV with study-aware folds (excluding transition)
set.seed(123)

folds <- createFolds(y_stage, k = 10, list = TRUE)
acc_vec <- numeric(length(folds))
kappa_vec <- numeric(length(folds))

for (i in seq_along(folds)) {

  test_idx  <- folds[[i]]
  train_idx <- setdiff(seq_len(nrow(X50_stage)), test_idx)

  rf_cv <- randomForest(
    x = X50_stage[train_idx, , drop = FALSE],
    y = y_stage[train_idx],
    ntree = 500
  )

  pred <- predict(rf_cv, X50_stage[test_idx, , drop = FALSE])
  cm <- confusionMatrix(pred, y_stage[test_idx])

  acc_vec[i] <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[i] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)


```




#dot plots with error bars for CV mean ± SD and test set performance for both splits
```{r}
library(ggplot2)

dev_comp <- data.frame(
  Split = c("SMART", "NAIVE"),

  # Test-set (confusionMatrix overall)
  test_acc   = c(0.7967, 0.9920),
  test_kappa = c(0.6145, 0.9855),

  # 10-fold CV (mean ± sd)
  cv_acc     = c(0.8661891, 0.9933033),
  cv_acc_sd  = c(0.1043786, 0.003840477),
  cv_kappa   = c(0.7410079, 0.9878995),
  cv_kappa_sd= c(0.1926055, 0.006933309)
)

dev_comp$Split <- factor(dev_comp$Split, levels = c("SMART", "NAIVE"))


```






```{r}
# Accuracy plot

p_acc <- ggplot(dev_comp, aes(x = Split)) +
  geom_errorbar(aes(ymin = cv_acc - cv_acc_sd, ymax = cv_acc + cv_acc_sd),
                width = 0.15, alpha = 0.8) +
  geom_point(aes(y = cv_acc, shape = "CV mean"), size = 3) +
  geom_point(aes(y = test_acc, shape = "Test"), size = 3) +
  ylim(0, 1) +
  labs(
    title = "Developmental stage (PC1–50): SMART vs NAIVE (Accuracy)",
    x = "Split strategy",
    y = "Accuracy",
    shape = ""
  ) +
  theme_minimal()

p_acc


```





```{r}
# Kappa plot

p_kappa <- ggplot(dev_comp, aes(x = Split)) +
  geom_errorbar(aes(ymin = cv_kappa - cv_kappa_sd, ymax = cv_kappa + cv_kappa_sd),
                width = 0.15, alpha = 0.8) +
  geom_point(aes(y = cv_kappa, shape = "CV mean"), size = 3) +
  geom_point(aes(y = test_kappa, shape = "Test"), size = 3) +
  ylim(0, 1) +
  labs(
    title = "Developmental stage (PC1–50): SMART vs NAIVE (Kappa)",
    x = "Split strategy",
    y = "Kappa",
    shape = ""
  ) +
  theme_minimal()

p_kappa


```




```{r}
# load csv file
metadata_all <- read.csv("metadata_with_treatment_predictions.csv")

```



```{r}
print(head(metadata_all))
```





**final model training and deployment to predict unknown samples with 50 PCs **


```{r}
# here i loaded my metadata with included treatment predictions 
meta2 <- read.csv("metadata_with_treatment_predictions.csv", stringsAsFactors = FALSE)

```



```{r}
#rename
meta_all <- metadata_all 

```



```{r}
# 1) Set the label column name for developmental stage
label_col <- "developmental_stage"

```




```{r}
# 2) Build PC matrix for ALL samples (must match meta row order used for PCA)
X50_all <- as.data.frame(pca_res$x[, 1:50, drop = FALSE])
stopifnot(nrow(X50_all) == nrow(meta_all))

```




```{r}
# 3) Define known vs unknown (unknown = NA or empty string)
known_idx   <- !is.na(meta_all[[label_col]]) & meta_all[[label_col]] != ""
unknown_idx <-  is.na(meta_all[[label_col]]) | meta_all[[label_col]] == ""

```




```{r}
#sanity checks
cat("Known samples:", sum(known_idx), "\n")
cat("Unknown samples:", sum(unknown_idx), "\n")
cat("Total samples:", nrow(meta_all), "\n")

```





```{r}
# 4) Train FINAL RF on ALL known samples
X_train_all <- X50_all[known_idx, , drop = FALSE]
y_train_all <- droplevels(factor(meta_all[[label_col]][known_idx]))

table(y_train_all)  # must show >=2 classes (otherwise RF will error)

set.seed(123)
rf_final_dev <- randomForest(
  x = X_train_all,
  y = y_train_all,
  ntree = 500,
  classwt = 1 / table(y_train_all)
)


```



```{r}
# 5) Predict unknown samples (+ probabilities + confidence)
X_unknown <- X50_all[unknown_idx, , drop = FALSE]

prob_unknown <- predict(rf_final_dev, X_unknown, type = "prob")
pred_unknown <- colnames(prob_unknown)[max.col(prob_unknown)]
conf_unknown <- apply(prob_unknown, 1, max)


```




```{r}
# 6) Quick sanity checks (distribution + confidence)
length(pred_unknown)
prop.table(table(pred_unknown))

```



```{r}
#check if model is just predicting mostly one class for unknown samples (which could indicate low confidence or bias)
table(pred_unknown)

```



```{r}
## confidence summary (max prob per sample)
# this gives an idea of how confident the model is in its predictions for the unknown samples (higher max prob = more confident)
# in simple terms, this shows the distribution of the highest predicted probability across all unknown samples, which can indicate how strongly the model is leaning towards its predicted class for those samples 

# # check confidence distribution quantiles, in simple terms, this shows the range of confidence levels for the model's predictions on the unknown samples, giving insight into how certain the model is across different samples (e.g., 10% of samples have max prob below 0.6, 50% below 0.8, etc.) 

summary(conf_unknown)

quantile(conf_unknown, c(0.1, 0.25, 0.5, 0.75, 0.9))

```



```{r}
# # here we can also check the agreement of predictions across multiple runs to see if the model is stable in its predictions for the unknown samples (since we don't have true labels, this is a way to assess consistency)

pred_list <- vector("list", 5)

for (s in 1:5) {
  set.seed(100 + s)

  rf_tmp <- randomForest(
    x = X_train_all,
    y = y_train_all,
    ntree = 500,
    classwt = 1 / table(y_train_all)
  )

  pred_list[[s]] <- as.character(predict(rf_tmp, X_unknown))
}

agree_matrix <- do.call(cbind, pred_list)

# agreement rate
agree_rate <- mean(apply(agree_matrix, 1, function(x) length(unique(x)) == 1))
agree_rate


```


```{r}
# # 7) Write predictions back into metadata (new columns)
# meta_all$dev_stage_pred <- NA_character_
# meta_all$dev_stage_conf <- NA_real_
# 
# meta_all$dev_stage_pred[unknown_idx] <- pred_unknown
# meta_all$dev_stage_conf[unknown_idx] <- conf_unknown
# 
# # also keep "final label" column filled
# meta_all$developmental_stage_filled <- meta_all[[label_col]]
# meta_all$developmental_stage_filled[unknown_idx] <- pred_unknown
# 
# # 8) Save updated metadata (now includes treatment + dev-stage)
# write.csv(meta_all,
#           "metadata_with_treatment_and_devstage_predictions.csv",
#           row.names = FALSE)


```




**final model training and deployment to predict unknown samples with 100 PCs **

```{r}
# 2) Build PC matrix for ALL samples (must match meta row order used for PCA)
X100_all <- as.data.frame(pca_res$x[, 1:100, drop = FALSE])
stopifnot(nrow(X50_all) == nrow(meta_all))

```



```{r}
# 3) Define known vs unknown (unknown = NA or empty string)
known_idx   <- !is.na(meta_all[[label_col]]) & meta_all[[label_col]] != ""
unknown_idx <-  is.na(meta_all[[label_col]]) | meta_all[[label_col]] == ""

```



```{r}
#sanity checks
cat("Known samples:", sum(known_idx), "\n")
cat("Unknown samples:", sum(unknown_idx), "\n")
cat("Total samples:", nrow(meta_all), "\n")

```




```{r}
# 4) Train FINAL RF on ALL known samples
X_train_all <- X100_all[known_idx, , drop = FALSE]
y_train_all <- droplevels(factor(meta_all[[label_col]][known_idx]))

table(y_train_all)  # must show >=2 classes (otherwise RF will error)

set.seed(123)
rf_final_dev <- randomForest(
  x = X_train_all,
  y = y_train_all,
  ntree = 500,
  classwt = 1 / table(y_train_all)
)

```




```{r}
# 5) Predict unknown samples (+ probabilities + confidence)
X_unknown <- X100_all[unknown_idx, , drop = FALSE]

prob_unknown <- predict(rf_final_dev, X_unknown, type = "prob")
pred_unknown <- colnames(prob_unknown)[max.col(prob_unknown)]
conf_unknown <- apply(prob_unknown, 1, max)


```




```{r}
# 6) Quick sanity checks (distribution + confidence)
length(pred_unknown)
prop.table(table(pred_unknown))

```




```{r}
#check if model is just predicting mostly one class for unknown samples (which could indicate low confidence or bias)
table(pred_unknown)

```



```{r}
## confidence summary (max prob per sample)
# this gives an idea of how confident the model is in its predictions for the unknown samples (higher max prob = more confident)
# in simple terms, this shows the distribution of the highest predicted probability across all unknown samples, which can indicate how strongly the model is leaning towards its predicted class for those samples 

# # check confidence distribution quantiles, in simple terms, this shows the range of confidence levels for the model's predictions on the unknown samples, giving insight into how certain the model is across different samples (e.g., 10% of samples have max prob below 0.6, 50% below 0.8, etc.) 

summary(conf_unknown)

quantile(conf_unknown, c(0.1, 0.25, 0.5, 0.75, 0.9))

```



```{r}
# # here we can also check the agreement of predictions across multiple runs to see if the model is stable in its predictions for the unknown samples (since we don't have true labels, this is a way to assess consistency)

pred_list <- vector("list", 5)

for (s in 1:5) {
  set.seed(100 + s)

  rf_tmp <- randomForest(
    x = X_train_all,
    y = y_train_all,
    ntree = 500,
    classwt = 1 / table(y_train_all)
  )

  pred_list[[s]] <- as.character(predict(rf_tmp, X_unknown))
}

agree_matrix <- do.call(cbind, pred_list)

# agreement rate
agree_rate <- mean(apply(agree_matrix, 1, function(x) length(unique(x)) == 1))
agree_rate

```


```{r}
```


```{r}
```

