

```{r}
getwd()


```



```{r}
#set working directory
setwd("C:/Users/yarram/Documents/Rna-seq-metadata-gap-filling/barley_rna_seq_data")



```




```{r}
#load necessary libraries
#loading libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(umap)
library(sva)
library(limma)
library(factoextra)
library(pheatmap)


```





```{r}
# read in the four parts of the  data
df <- read.delim("../barley_rna_seq_data/my_barley_input.2.part1.abundance.tsv.gz", header = TRUE, sep = "\t")
df1 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part2.abundance.tsv.gz", header = TRUE, sep = "\t")
df2 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part3.abundance.tsv.gz", header = TRUE, sep = "\t")
df3 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part4.abundance.tsv.gz", header = TRUE, sep = "\t")




```





```{r}

# Rename the first column of each df to "gene"
names(df)[1]   <- "gene"
names(df1)[1]  <- "gene"
names(df2)[1]  <- "gene"
names( df3)[1] <- "gene"


```





```{r}

library(dplyr)
# Merge data frames by "gene" column using full_join
df4 <- list(df, df1, df2, df3) |>
  Reduce(function(x, y) full_join(x, y, by = "gene"), x = _)

# (optional) if any duplicate sample names exist, make them unique
colnames(df4) <- make.unique(colnames(df4))



```






```{r}
sum(is.na(df4)) 


```





```{r}
# extract TPM columns + gene column
tpm_df <- df4 %>%
  select(gene, ends_with("_tpm"))


```





```{r}
# quick check
dim(tpm_df)
head(tpm_df[, 1:6])


```





```{r}
#set row names as gene names and remove gene column

rownames(tpm_df) <- tpm_df$gene
tpm_df <- tpm_df[, -1]
#convert to matrix
tpm_matrix <- as.matrix(tpm_df)
mode(tpm_matrix) <- "numeric"

```





```{r}
#check dimensions
dim(tpm_matrix)


```





```{r}

#check first few samples
head(tpm_matrix[, 1:6])
head(tpm_matrix[, 1:6])


```





```{r}

#remove _tpm from column names so that they match metadata
colnames(tpm_matrix) <- gsub("_tpm$", "", colnames(tpm_matrix))


```





```{r}
#load your metadata
library(readxl)
meta_raw <- read_xlsx("hordeum_vulgare_10520.xlsx")



```





```{r}
# what are column names
colnames(meta_raw)


```






```{r}

# i want to see how many sheets of data are in the excel file
excel_sheets("hordeum_vulgare_10520.xlsx")

```






```{r}
#load necessary libraries before data cleaning
library(dplyr)
library(magrittr)  # makes sure %>% is available

```





```{r}
#i want to read in a specific sheet from the excel file
meta_raw <- read_xlsx("hordeum_vulgare_10520.xlsx", sheet = "Sheet1")


```





```{r}

#clean/standardize metadata (use your real columns)
meta <- meta_raw %>%
  rename(
    sample_id = Run,     # SRR/DRR/ERR
    tissue    = sample   # your tissue column is named sample
  ) %>%
  mutate(
    sample_id = as.character(sample_id),
    treatment = trimws(as.character(treatment)),
    layout    = tolower(trimws(as.character(layout))),
    tissue    = tolower(trimws(as.character(tissue)))
  ) %>%
  distinct(sample_id, .keep_all = TRUE)

```





```{r}
# checking the data
dim(meta)
head(meta[, c("sample_id","treatment","tissue","layout")])

```





```{r}
# keep only samples that are present in the TPM matrix, one row per sample_id
meta_keep <- meta %>%
  filter(sample_id %in% colnames(tpm_matrix)) %>%
  distinct(sample_id, .keep_all = TRUE)


```





```{r}
# order rows of metadata to match the TPM column order
ord  <- match(colnames(tpm_matrix), meta_keep$sample_id)
keep <- !is.na(ord)


```





```{r}
# apply the same filter/order to both objects
tpm_matrix   <- tpm_matrix[, keep, drop = FALSE]
meta_aligned <- meta_keep[ord[keep], ]


```





```{r}
# sanity checks
ncol(tpm_matrix)            # should equal:
nrow(meta_aligned)          # ...this
sum(is.na(meta_aligned$sample_id))  # should be 0
head(meta_aligned[, c("sample_id","treatment","tissue","layout")])


```





```{r}
# log1p-transform and transpose -> rows = samples, cols = genes
# RNA-seq TPM values can vary hugely — some genes have TPMs of 0.1, others 10,000+.
#Without transforming, the few highly expressed genes dominate the analysis.
#It stabilizes variance (so big and small expression values are comparable).

log_tpm <- t(log1p(tpm_matrix))   # big object: 10096 x ~37962


```







```{r}

# remove genes with zero variance (constant across all samples)
# This finds genes that have no change across samples — i.e., their expression is constant everywhere.
# Removing zero-variance genes keeps only the informative ones — genes that actually differ between samples — improving PCA/UMAP clarity and speed.
gene_var <- apply(log_tpm, 2, var)
sum_zero_var <- sum(gene_var == 0)         # how many genes removed?
log_tpm_f <- log_tpm[, gene_var > 0, drop = FALSE]


```







```{r}
#  quick checks
ncol_before <- ncol(log_tpm)
ncol_after  <- ncol(log_tpm_f)
cat("genes before:", ncol_before, "\ngenes after:", ncol_after,
    "\nzero-variance genes removed:", sum_zero_var, "\n")


```






```{r}
# run PCA on log-transformed TPM
pca_res <- prcomp(log_tpm_f, center = TRUE, scale. = TRUE)

```






```{r}
# variance explained by first few PCs
pc_var <- summary(pca_res)$importance[2, 1:6] * 100
pc_var

```







```{r}
#prepare the data for random forest 10 PCs,only known tissue and known SRA study)

# 1) Make a clean metadata object with a safe sra_study column

#in this step, we are creating a new metadata dataframe called meta2 from the existing meta_aligned dataframe.

#why is this step important?
# This step is crucial because it standardizes and prepares the metadata for subsequent analysis.
# By creating a new column sra_study and ensuring that the tissue column is in character format,
# we make it easier to filter and manipulate the data later on, especially when we want
# to select samples with known tissue types and study identifiers for training/testing our model.


library(dplyr)


meta2 <- meta_aligned %>%
  mutate(
    sra_study = `SRA Study`,   # column name has a SPACE in your file
    tissue = as.character(tissue)
  )


```






```{r}
# 2) Keep only samples that have BOTH tissue + sra_study (for training/testing)

# In this step, we are filtering the metadata dataframe meta2 to retain only those samples that have known values for both the tissue and sra_study columns.
# This is important because we want to ensure that our training and testing datasets only include samples with complete information for these key variables.
# By doing this, we can build a more reliable model since we won't be dealing with missing data for these critical features.


keep_known <- !is.na(meta2$tissue) & meta2$tissue != "" & !is.na(meta2$sra_study) & meta2$sra_study != ""

meta_known <- meta2[keep_known, , drop = FALSE]




```





```{r}
#3) Build features from first 10 PCs (rows match samples)
# In this step, we are creating a new feature matrix X10 that contains the first 10 principal components (PCs) from the PCA results for only those samples that have known tissue types.
# We are also creating a response vector y that contains the corresponding tissue labels for these samples.
# why is this step important?
# This step is crucial because it transforms the high-dimensional gene expression data into a more manageable form
# by using the first 10 principal components (PCs) as features.
# These PCs capture the most significant patterns of variation in the data,
# allowing us to reduce noise and focus on the most informative aspects for classification.
# By selecting only the samples with known tissue types (as determined in the previous step),
# we ensure that our feature set aligns with the labels we want to predict,
# which is essential for training an effective machine learning model.

X10 <- as.data.frame(pca_res$x[keep_known, 1:10, drop = FALSE])
y  <- factor(meta_known$tissue)



```





```{r}
# Quick sanity checks
dim(X10)                 # should be (num_known_samples x 10)
table(y)                 # counts per tissue in training set
length(unique(meta_known$sra_study))  # how many studies used

```




```{r}
# 3) now i am doing study -wise splitting so that training and testing sets do not share samples from same study


#why i did this step?

# This step is important because it ensures that the training and testing datasets are independent in terms of study origin.
# By splitting the data based on unique sra_study identifiers, we prevent data leakage,
# where samples from the same study could end up in both the training and testing sets.
# Such leakage can lead to overly optimistic performance estimates, as the model might learn study-specific patterns
# rather than generalizable features.
# By keeping studies separate, we can better assess how well the model generalizes to new, unseen studies,
# which is crucial for evaluating its real-world applicability.


set.seed(123)

studies <- unique(meta_known$sra_study)
test_studies <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_known$sra_study %in% train_studies
test_idx  <- meta_known$sra_study %in% test_studies

X_train <- X10[train_idx, , drop = FALSE]
y_train <- y[train_idx]

X_test  <- X10[test_idx, , drop = FALSE]
y_test  <- y[test_idx]

dim(X_train); dim(X_test)
length(unique(meta_known$sra_study[train_idx]))
length(unique(meta_known$sra_study[test_idx]))



```






```{r}
# final sanity checks

dim(X_train); dim(X_test)
length(unique(meta_known$sra_study[train_idx]))
length(unique(meta_known$sra_study[test_idx]))



```

# i splitted known samples for training and testing based on SRA study so that no overlap of studies in training and testing

#5179 are samples in training and 10 is PCs used for features

#684 are samples in testing and 10 is PCs used for features

#156 unique sra studies in training set
#39 unique sra studies in testing set

#“I split by SRA study: 156 studies for training, 39 for testing (no overlap)”

# why 5179 plus 684 = 5863 known samples and each represented and by 10 PCs
#156 SRA studies are UNIQUE studies, and each study contains many samples
#those 156 SRA studies are unique, and the same study appears many times in your 5179 samples.





```{r}
# perform random forest classification

#want to install randomForest package and carpet package if not already installed
#install.packages("randomForest")
#install.packages("caret")

#load libraries
library(randomForest)
library(caret)


set.seed(123)

# train RF

#this code is training a random forest classifier using the training data (X_train and y_train).
# The randomForest function is called with the following parameters:
# x = X_train: This is the feature matrix for training, where each row represents a sample and each column represents a feature (in this case, the first 10 principal components).
# y = as.factor(y_train): This is the response vector for training, which contains the class labels (tissue types) for each sample. The labels are converted to a factor to indicate that they are categorical.
# ntree = 500: This specifies the number of trees to grow in the random forest. A higher number of trees can improve model performance but also increases computation time.
# importance = TRUE: This option indicates that the model should calculate and store variable importance measures, which can be useful for understanding which features contribute most to the classification.


rf10 <- randomForest(
  x = X_train,
  y = as.factor(y_train),
  ntree = 500,
  importance = TRUE
)

# predict on TEST

#this code is using the trained random forest model (rf10) to make predictions on the test dataset (X_test).
# The predict function is called with the following parameters:
# rf10: This is the trained random forest model that was created in the previous step.
# X_test: This is the feature matrix for the test dataset, where each row represents a sample and each column represents a feature (the first 10 principal components).
# The function returns the predicted class labels for each sample in the test set, which are stored in the variable pred_test.

pred_test <- predict(rf10, X_test)

# confusion matrix + metrics
#this code is generating a confusion matrix to evaluate the performance of the random forest model's predictions on the test dataset.
# The confusionMatrix function from the caret package is called with the following parameters:
# pred_test: This is the vector of predicted class labels obtained from the random forest model for the test dataset.
# as.factor(y_test): This is the true class labels for the test dataset, converted to a factor to indicate that they are categorical.
# The confusion matrix provides a summary of the model's performance by comparing the predicted labels to the true labels,
# allowing for the calculation of various metrics such as accuracy, sensitivity, specificity, and more.

cm <- confusionMatrix(pred_test, as.factor(y_test))
cm



```





```{r}
# view predicted probabilities for test set

#this code is using the trained random forest model (rf10) to obtain the predicted class probabilities for each sample in the test dataset (X_test).  
# The predict function is called with the following parameters:
# rf10: This is the trained random forest model created earlier.
# X_test: This is the feature matrix for the test dataset, where each row represents a
# sample and each column represents a feature (the first 10 principal components).
# type = "prob": This option specifies that the function should return the predicted probabilities for each
# class rather than the predicted class labels.
# The resulting prob_test object contains the predicted probabilities for each class (tissue type) for each sample in the test set.  


prob_test <- predict(rf10, X_test, type = "prob")
head(prob_test)



```





```{r}
#next step is 10 fold cross validation


set.seed(123)

unique_studies <- unique(meta_known$sra_study)

# randomly assign each study to one of 10 folds
fold_id <- sample(rep(1:10, length.out = length(unique_studies)))

study_folds <- data.frame(
  sra_study = unique_studies,
  fold = fold_id
)


```




```{r}
#storage for results

library(randomForest)
library(caret)

acc_vec <- numeric(10)
kappa_vec <- numeric(10)


```






```{r}
#now perform 10 fold cross validation by loop

for (k in 1:10) {

  test_studies <- study_folds$sra_study[study_folds$fold == k]
  
  test_idx  <- meta_known$sra_study %in% test_studies
  train_idx <- !test_idx
  
  X_train <- X10[train_idx, ]
  y_train <- y[train_idx]
  
  X_test <- X10[test_idx, ]
  y_test <- y[test_idx]
  
  rf <- randomForest(
    x = X_train,
    y = y_train,
    ntree = 500
  )
  
  pred <- predict(rf, X_test)
  
  cm <- confusionMatrix(pred, y_test)
  
  acc_vec[k] <- cm$overall["Accuracy"]
  kappa_vec[k] <- cm$overall["Kappa"]
}



```







```{r}
# summrisize cross-validation results

mean(acc_vec)
sd(acc_vec)

mean(kappa_vec)
sd(kappa_vec)



```





```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```

