
 
```{r}
getwd()

```





```{r}
#set working directory
setwd("C:/Users/yarram/Documents/Rna-seq-metadata-gap-filling/barley_rna_seq_data")

```





```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(umap)
library(sva)
library(limma)
library(factoextra)
library(pheatmap)

```





```{r}
# read in the four parts of the  data
df <- read.delim("../barley_rna_seq_data/my_barley_input.2.part1.abundance.tsv.gz", header = TRUE, sep = "\t")
df1 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part2.abundance.tsv.gz", header = TRUE, sep = "\t")
df2 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part3.abundance.tsv.gz", header = TRUE, sep = "\t")
df3 <- read.delim("../barley_rna_seq_data/my_barley_input.2.part4.abundance.tsv.gz", header = TRUE, sep = "\t")


```






```{r}
# Rename the first column of each df to "gene"
names(df)[1]   <- "gene"
names(df1)[1]  <- "gene"
names(df2)[1]  <- "gene"
names( df3)[1] <- "gene"

```





```{r}

library(dplyr)
# Merge data frames by "gene" column using full_join
df4 <- list(df, df1, df2, df3) |>
  Reduce(function(x, y) full_join(x, y, by = "gene"), x = _)

# (optional) if any duplicate sample names exist, make them unique
colnames(df4) <- make.unique(colnames(df4))


```






```{r}
# check for missing values
sum(is.na(df4)) 

```






```{r}
# extract TPM columns + gene column
tpm_df <- df4 %>%
  select(gene, ends_with("_tpm")) 


```






```{r}
# quick check
dim(tpm_df)
head(tpm_df[, 1:6])


```






```{r}
#set row names as gene names and remove gene column

rownames(tpm_df) <- tpm_df$gene
tpm_df <- tpm_df[, -1]
#convert to matrix
tpm_matrix <- as.matrix(tpm_df)
mode(tpm_matrix) <- "numeric"


```





```{r}
#check dimensions
dim(tpm_matrix)

```




```{r}
#check first few samples
head(tpm_matrix[, 1:6])
head(tpm_matrix[, 1:6])

```






```{r}
#remove _tpm from column names so that they match metadata
colnames(tpm_matrix) <- gsub("_tpm$", "", colnames(tpm_matrix))

```





```{r}
#load your metadata
library(readxl)
meta_raw <- read_xlsx("hordeum_vulgare_10520.xlsx")

```





```{r}
# what are column names
colnames(meta_raw)


```





```{r}
# i want to see how many sheets of data are in the excel file
excel_sheets("hordeum_vulgare_10520.xlsx")

```






```{r}
#load necessary libraries before data cleaning
library(dplyr)
library(magrittr)  # makes sure %>% is available


```





```{r}
#i want to read in a specific sheet from the excel file
meta_raw <- read_xlsx("hordeum_vulgare_10520.xlsx", sheet = "Sheet1")

```






```{r}

#clean/standardize metadata (use your real columns)
meta <- meta_raw %>%
  rename(
    sample_id = Run,     # SRR/DRR/ERR
    tissue    = sample   # your tissue column is named sample
  ) %>%
  mutate(
    sample_id = as.character(sample_id),
    treatment = trimws(as.character(treatment)),
    layout    = tolower(trimws(as.character(layout))),
    tissue    = tolower(trimws(as.character(tissue)))
  ) %>%
  distinct(sample_id, .keep_all = TRUE)



```






```{r}
# checking the data
dim(meta)
head(meta[, c("sample_id","treatment","tissue","layout")])


```





```{r}
# keep only samples that are present in the TPM matrix, one row per sample_id
meta_keep <- meta %>%
  filter(sample_id %in% colnames(tpm_matrix)) %>%
  distinct(sample_id, .keep_all = TRUE)


```





```{r}
# keep only samples that are present in the TPM matrix, one row per sample_id
meta_keep <- meta %>%
  filter(sample_id %in% colnames(tpm_matrix)) %>%
  distinct(sample_id, .keep_all = TRUE)


```





```{r}
# order rows of metadata to match the TPM column order
ord  <- match(colnames(tpm_matrix), meta_keep$sample_id)
keep <- !is.na(ord)


```






```{r}
# apply the same filter/order to both objects
tpm_matrix   <- tpm_matrix[, keep, drop = FALSE]
meta_aligned <- meta_keep[ord[keep], ]


```






```{r}

# sanity checks
ncol(tpm_matrix)            # should equal:
nrow(meta_aligned)          # ...this
sum(is.na(meta_aligned$sample_id))  # should be 0
head(meta_aligned[, c("sample_id","treatment","tissue","layout")])

```







```{r}
# log1p-transform and transpose -> rows = samples, cols = genes
# RNA-seq TPM values can vary hugely — some genes have TPMs of 0.1, others 10,000+.
#Without transforming, the few highly expressed genes dominate the analysis.
#It stabilizes variance (so big and small expression values are comparable).

log_tpm <- t(log1p(tpm_matrix))   # big object: 10096 x ~37962


```





```{r}
# remove genes with zero variance (constant across all samples)
# This finds genes that have no change across samples — i.e., their expression is constant everywhere.
# Removing zero-variance genes keeps only the informative ones — genes that actually differ between samples — improving PCA/UMAP clarity and speed.
gene_var <- apply(log_tpm, 2, var)
sum_zero_var <- sum(gene_var == 0)         # how many genes removed?
log_tpm_f <- log_tpm[, gene_var > 0, drop = FALSE]


```





```{r}
#  quick checks
ncol_before <- ncol(log_tpm)
ncol_after  <- ncol(log_tpm_f)
cat("genes before:", ncol_before, "\ngenes after:", ncol_after,
    "\nzero-variance genes removed:", sum_zero_var, "\n")



```






```{r}
# run PCA on log-transformed TPM
pca_res <- prcomp(log_tpm_f, center = TRUE, scale. = TRUE)

```






```{r}
# variance explained by first few PCs
pc_var <- summary(pca_res)$importance[2, 1:6] * 100
pc_var


```






```{r}
#prepare data random forest 10 Pcs,only for know treatment

library(dplyr)
meta2 <- meta_aligned %>%
  mutate(
    sra_study = `SRA Study`,   # column name has a SPACE in your file
    treatment = as.character(treatment)
  )

```





```{r}
# keep only samples with known treatment and SRA study
keep_known <- !is.na(meta2$treatment) & meta2$treatment != "" & !is.na(meta2$sra_study) & meta2$sra_study != ""
meta_known <- meta2[keep_known, , drop = FALSE]


```





```{r}
# prepare data for random forest:

X10 <- as.data.frame(pca_res$x[keep_known, 1:10, drop = FALSE])
y  <- factor(meta_known$treatment)

```





```{r}
# sanity checks
dim(meta_known)
table(meta_known$treatment)
length(unique(meta_known$sra_study))

```



```{r}
# group into fewer classes (edit mapping if you want)
meta_known <- meta_known %>%
  mutate(
    treatment_group = case_when(
      treatment == "control" ~ "control",

      treatment %in% c("abiotic stress","drought","salt","heat","cold","oxidative stress") ~ "abiotic",
      treatment %in% c("pathogen","symbiotic") ~ "biotic",

      treatment %in% c("hormone treatment") ~ "hormone",
      treatment %in% c("nutrient") ~ "nutrient",

      treatment %in% c("biostimulant") ~ "biostimulant",
      treatment %in% c("combined(sym/bio)") ~ "sym/bio",
      treatment %in% c("genetic modification") ~ "geneticmodification",

      TRUE ~ "other"
    )
  )

```





```{r}
# check class distribution
table(meta_known$treatment_group)

```



```{r}
# for random forest, we want to focus on the main classes with enough samples. Adjust as needed.
keep_main <- meta_known$treatment_group %in% c("control","abiotic","biotic")
meta_treat <- meta_known[keep_main, , drop = FALSE]
y <- droplevels(factor(meta_treat$treatment_group))
table(y)

```



```{r}
# prepare X matrix for random forest, keeping only samples with known treatment and in main classes
X10 <- as.data.frame(pca_res$x[keep_known, 1:10, drop = FALSE])
X10 <- X10[keep_main, , drop = FALSE]


```




```{r}
# sanity checks
table(y)
dim(X10)
length(unique(meta_treat$sra_study))

```





```{r}
# split into training and test sets, ensuring samples from the same SRA study are in the same set
set.seed(123)

studies <- unique(meta_treat$sra_study)

test_studies  <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_treat$sra_study %in% train_studies
test_idx  <- meta_treat$sra_study %in% test_studies

X_train <- X10[train_idx, , drop = FALSE]
y_train <- droplevels(y[train_idx])

X_test  <- X10[test_idx, , drop = FALSE]
y_test  <- factor(y[test_idx], levels = levels(y_train))

dim(X_train); dim(X_test)
table(y_train)



```




```{r}
# train random forestlibrary(randomForest)

library(caret)

set.seed(123)

rf_treat <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
)

pred_test <- predict(rf_treat, X_test)
cm_treat  <- confusionMatrix(pred_test, y_test)

cm_treat


```




```{r}
# probabilitis
prob_test10 <- predict(rf_treat, X_test, type = "prob")
head(prob_test10)

```




```{r}
# predicted class from probabilities
pred_from_prob <- colnames(prob_test10)[max.col(prob_test10)]
pred_from_prob <- factor(pred_from_prob, levels = levels(y_train))


```





```{r}
#cv loop

library(randomForest)
library(caret)

set.seed(123)

studies <- unique(meta_treat$sra_study)
folds   <- createFolds(studies, k = 10, list = TRUE)

acc_vec   <- numeric(length(folds))
kappa_vec <- numeric(length(folds))

for (k in 1:10) {

  test_studies <- studies[folds[[k]]]

  test_idx  <- meta_treat$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X10[train_idx, , drop = FALSE]
  y_train <- y[train_idx]

  X_test  <- X10[test_idx, , drop = FALSE]
  y_test  <- y[test_idx]

 rf_cv <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  classwt = 1 / table(y_train)
)

  pred <- predict(rf_cv, X_test, type = "response")
  cm   <- confusionMatrix(pred, y_test)

  acc_vec[k]   <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)

```



```{r}
# after adding the class weights, the class distribution
prop.table(table(y))

```



#with 20 PCs

```{r}
# # prepare X matrix for random forest, keeping only samples with known treatment and in main classes
X20 <- as.data.frame(pca_res$x[keep_known, 1:20, drop = FALSE])
X20 <- X20[keep_main, , drop = FALSE]


```




```{r}
# sanity checks
table(y)
dim(X20)
length(unique(meta_treat$sra_study))

```


```{r}
# group into fewer classes (edit mapping if you want)
meta_known <- meta_known %>%
  mutate(
    treatment_group = case_when(
      treatment == "control" ~ "control",

      treatment %in% c("abiotic stress","drought","salt","heat","cold","oxidative stress") ~ "abiotic",
      treatment %in% c("pathogen","symbiotic") ~ "biotic",

      treatment %in% c("hormone treatment") ~ "hormone",
      treatment %in% c("nutrient") ~ "nutrient",

      treatment %in% c("biostimulant") ~ "biostimulant",
      treatment %in% c("combined(sym/bio)") ~ "sym/bio",
      treatment %in% c("genetic modification") ~ "geneticmodification",

      TRUE ~ "other"
    )
  )

```




```{r}
# check class distribution
table(meta_known$treatment_group)

```



```{r}
# for random forest, we want to focus on the main classes with enough samples. Adjust as needed.
keep_main <- meta_known$treatment_group %in% c("control","abiotic","biotic")
meta_treat <- meta_known[keep_main, , drop = FALSE]
y <- droplevels(factor(meta_treat$treatment_group))

table(y)

```



```{r}
# prepare X matrix for random forest, keeping only samples with known treatment and in main classes
X20 <- as.data.frame(pca_res$x[keep_known, 1:20, drop = FALSE])
X20 <- X20[keep_main, , drop = FALSE]

```




```{r}
# sanity checks
table(y)
dim(X20)
length(unique(meta_treat$sra_study))

```


```{r}
# split into training and test sets, ensuring samples from the same SRA study are in the same set
set.seed(123)

studies <- unique(meta_treat$sra_study)

test_studies  <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_treat$sra_study %in% train_studies
test_idx  <- meta_treat$sra_study %in% test_studies

X_train <- X20[train_idx, , drop = FALSE]
y_train <- droplevels(y[train_idx])

X_test  <- X20[test_idx, , drop = FALSE]
y_test  <- factor(y[test_idx], levels = levels(y_train))

dim(X_train); dim(X_test)
table(y_train)

```


```{r}
# train random forestlibrary(randomForest)

library(caret)

set.seed(123)

rf_treat <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
)

pred_test <- predict(rf_treat, X_test)
cm_treat  <- confusionMatrix(pred_test, y_test)

cm_treat

```


```{r}
# probabilitis
prob_test20 <- predict(rf_treat, X_test, type = "prob")
head(prob_test20)

```


```{r}
#cv loop

library(randomForest)
library(caret)

set.seed(123)

studies <- unique(meta_treat$sra_study)
folds   <- createFolds(studies, k = 10, list = TRUE)

acc_vec   <- numeric(length(folds))
kappa_vec <- numeric(length(folds))

for (k in 1:10) {

  test_studies <- studies[folds[[k]]]

  test_idx  <- meta_treat$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X20[train_idx, , drop = FALSE]
  y_train <- y[train_idx]

  X_test  <- X20[test_idx, , drop = FALSE]
  y_test  <- y[test_idx]

 rf_cv <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  classwt = 1 / table(y_train)
)

  pred <- predict(rf_cv, X_test, type = "response")
  cm   <- confusionMatrix(pred, y_test)

  acc_vec[k]   <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)


```


#with 50 PCs

```{r}
# # prepare X matrix for random forest, keeping only samples with known treatment and in main classes
X50 <- as.data.frame(pca_res$x[keep_known, 1:50, drop = FALSE])
X50 <- X50[keep_main, , drop = FALSE]


```



```{r}
# sanity checks
table(y)
dim(X50)
length(unique(meta_treat$sra_study))

```


```{r}
# split into training and test sets, ensuring samples from the same SRA study are in the same set
set.seed(123)

studies <- unique(meta_treat$sra_study)

test_studies  <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_treat$sra_study %in% train_studies
test_idx  <- meta_treat$sra_study %in% test_studies

X_train <- X50[train_idx, , drop = FALSE]
y_train <- droplevels(y[train_idx])

X_test  <- X50[test_idx, , drop = FALSE]
y_test  <- factor(y[test_idx], levels = levels(y_train))

dim(X_train); dim(X_test)
table(y_train)

```



```{r}
# train random forestlibrary(randomForest)

library(caret)

set.seed(123)

rf_treat <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
)

pred_test <- predict(rf_treat, X_test)
cm_treat  <- confusionMatrix(pred_test, y_test)

cm_treat

```


```{r}
# probabilitis
prob_test50 <- predict(rf_treat, X_test, type = "prob")
head(prob_test50)

```


```{r}
#cv loop

library(randomForest)
library(caret)

set.seed(123)

studies <- unique(meta_treat$sra_study)
folds   <- createFolds(studies, k = 10, list = TRUE)

acc_vec   <- numeric(length(folds))
kappa_vec <- numeric(length(folds))

for (k in 1:10) {

  test_studies <- studies[folds[[k]]]

  test_idx  <- meta_treat$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X50[train_idx, , drop = FALSE]
  y_train <- y[train_idx]

  X_test  <- X50[test_idx, , drop = FALSE]
  y_test  <- y[test_idx]

 rf_cv <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  classwt = 1 / table(y_train)
)

  pred <- predict(rf_cv, X_test, type = "response")
  cm   <- confusionMatrix(pred, y_test)

  acc_vec[k]   <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)


```




# now i want to do this by extarcting 100 PCs
```{r}
# prepare data for random forest:
X100 <- as.data.frame(pca_res$x[keep_known, 1:100, drop = FALSE])
y  <- factor(meta_known$treatment)  

```




```{r}
# sanity checks
dim(meta_known)
table(meta_known$treatment)
length(unique(meta_known$sra_study))


```





```{r}
# check class distribution
table(meta_known$treatment_group)

```






```{r}
# for random forest, we want to focus on the main classes with enough samples. Adjust as needed.
keep_main <- meta_known$treatment_group %in% c("control","abiotic","biotic")
meta_treat <- meta_known[keep_main, , drop = FALSE]
y <- droplevels(factor(meta_treat$treatment_group))
table(y)


```






```{r}
# prepare X matrix for random forest, keeping only samples with known treatment and in main classes
X100 <- as.data.frame(pca_res$x[keep_known, 1:100, drop = FALSE])
X100 <- X100[keep_main, , drop = FALSE]


```






```{r}
# sanity checks
table(y)
dim(X100)
length(unique(meta_treat$sra_study))

```
```{r}
# split into training and test sets, ensuring samples from the same SRA study are in the same set
set.seed(123)
studies <- unique(meta_treat$sra_study)
test_studies  <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)
train_idx <- meta_treat$sra_study %in% train_studies
test_idx  <- meta_treat$sra_study %in% test_studies
X_train <- X100[train_idx, , drop = FALSE]
y_train <- droplevels(y[train_idx])

X_test  <- X100[test_idx, , drop = FALSE]
y_test  <- factor(y[test_idx], levels = levels(y_train))

dim(X_train); dim(X_test)
table(y_train)
```





```{r}
# train random forestlibrary(randomForest)
library(randomForest)
library(caret)
set.seed(123)
rf_treat_100 <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
)
pred_test_100 <- predict(rf_treat_100, X_test)
cm_treat_100  <- confusionMatrix(pred_test_100, y_test)
cm_treat_100


```






```{r}
# probabilitis
prob_test50 <- predict(rf_treat, X_test, type = "prob")
head(prob_test50)

```





```{r}
#cv loop

library(randomForest)
library(caret)

set.seed(123)

studies <- unique(meta_treat$sra_study)
folds   <- createFolds(studies, k = 10, list = TRUE)

acc_vec   <- numeric(length(folds))
kappa_vec <- numeric(length(folds))

for (k in 1:10) {

  test_studies <- studies[folds[[k]]]

  test_idx  <- meta_treat$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X100[train_idx, , drop = FALSE]
  y_train <- y[train_idx]

  X_test  <- X100[test_idx, , drop = FALSE]
  y_test  <- y[test_idx]

 rf_cv <- randomForest(
  x = X_train,
  y = y_train,
  ntree = 500,
  classwt = 1 / table(y_train)
)

  pred <- predict(rf_cv, X_test, type = "response")
  cm   <- confusionMatrix(pred, y_test)

  acc_vec[k]   <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)

```





#stupid -way for comparison
```{r}
library(randomForest)
library(caret)

set.seed(123)

# Create folds randomly at sample level (NOT study-aware)
folds <- createFolds(y, k = 10, list = TRUE)

acc_vec_stupid <- numeric(length(folds))
kappa_vec_stupid <- numeric(length(folds))

for (k in 1:10) {
  
  test_idx  <- folds[[k]]
  train_idx <- setdiff(seq_len(nrow(X100)), test_idx)
  
  X_train <- X100[train_idx, , drop = FALSE]
  y_train <- y[train_idx]
  
  X_test  <- X100[test_idx, , drop = FALSE]
  y_test  <- y[test_idx]
  
  rf_cv <- randomForest(
    x = X_train,
    y = y_train,
    ntree = 500,
    classwt = 1 / table(y_train)
  )
  
  pred <- predict(rf_cv, X_test, type = "response")
  cm   <- confusionMatrix(pred, y_test)
  
  acc_vec_stupid[k]   <- as.numeric(cm$overall["Accuracy"])
  kappa_vec_stupid[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec_stupid); sd(acc_vec_stupid)
mean(kappa_vec_stupid); sd(kappa_vec_stupid)

```





#dot plots with error bars for CV mean, ± SD and test set performance for both splits
```{r}
#creating data frame 

library(ggplot2)
summary_compare <- data.frame(
  split = c("SMART (study-aware)", "STUPID (sample-based)"),
  
  test_acc = c(0.4803, 0.9602338),
  test_kappa = c(0.2337, 0.9167351),
  
  cv_acc = c(0.4437874, 0.9602338),
  cv_acc_sd = c(0.1645924, 0.01551013),
  
  cv_kappa = c(0.07136937, 0.9167351),
  cv_kappa_sd = c(0.1197868, 0.03280107)
)

summary_compare

```






```{r}
# accuracy
p_acc <- ggplot(summary_compare, aes(x = split)) +
  geom_point(aes(y = cv_acc, shape = "CV mean"), size = 3) +
  geom_errorbar(aes(ymin = cv_acc - cv_acc_sd, ymax = cv_acc + cv_acc_sd), width = 0.15) +
  geom_point(aes(y = test_acc, shape = "Test"), size = 3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Treatment (PC1–100): SMART vs STUPID (Accuracy)",
    x = "Split strategy",
    y = "Accuracy",
    shape = ""
  ) +
  theme_minimal()

p_acc

```






```{r}
#kappa
p_kappa <- ggplot(summary_compare, aes(x = split)) +
  geom_point(aes(y = cv_kappa, shape = "CV mean"), size = 3) +
  geom_errorbar(aes(ymin = cv_kappa - cv_kappa_sd, ymax = cv_kappa + cv_kappa_sd), width = 0.15) +
  geom_point(aes(y = test_kappa, shape = "Test"), size = 3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Treatment (PC1–100): SMART vs STUPID (Kappa)",
    x = "Split strategy",
    y = "Kappa",
    shape = ""
  ) +
  theme_minimal()

p_kappa

```



#deeplearning (MLP)


```{r}
# installing deep learning packages
#install.packages("nnet")

```






```{r}
library(dplyr)
library(caret)
library(nnet)


```




```{r}
# sanity check
stopifnot(nrow(X100) == nrow(meta_treat))
stopifnot(length(y) == nrow(meta_treat))
table(y)
length(unique(meta_treat$sra_study))


```



```{r}
# split into train and test
set.seed(123)
studies <- unique(meta_treat$sra_study)
test_studies  <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_treat$sra_study %in% train_studies
test_idx  <- meta_treat$sra_study %in% test_studies

X_train <- X10[train_idx, , drop = FALSE]
X_test  <- X10[test_idx,  , drop = FALSE]

y_train <- droplevels(y[train_idx])
y_test  <- factor(y[test_idx], levels = levels(y_train))

dim(X_train); dim(X_test)
table(y_train); table(y_test)

```



```{r}
# Perform MLP

library(nnet)

set.seed(123)
mlp10 <- nnet(
  x = as.matrix(X_train),
  y = class.ind(y_train),
  size = 10,
  decay = 1e-3,
  maxit = 300,
  softmax = TRUE,
  trace = FALSE,
  MaxNWts = 100000
)

# --- 3) Predict + Confusion Matrix
prob_test <- predict(mlp10, as.matrix(X_test), type = "raw")
pred_test <- colnames(prob_test)[max.col(prob_test)]
pred_test <- factor(pred_test, levels = levels(y_train))   

cm_test <- confusionMatrix(pred_test, y_test)
cm_test

```



```{r}
# probabilities
prob_test <- predict(mlp10, as.matrix(X_test), type = "raw")
head(prob_test)

```



```{r}
# cross-validation MLP
library(caret)
library(nnet)

set.seed(123)

studies <- unique(meta_treat$sra_study)
folds <- createFolds(studies, k = 10, list = TRUE)

acc_vec   <- numeric(length(folds))
kappa_vec <- numeric(length(folds))

for (k in seq_along(folds)) {

  test_studies <- studies[folds[[k]]]
  test_idx  <- meta_treat$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X10[train_idx, , drop = FALSE]
  X_test  <- X10[test_idx,  , drop = FALSE]

  y_train <- droplevels(y[train_idx])
  y_test  <- factor(y[test_idx], levels = levels(y_train))  # align to train

  # remove test rows that became NA after alignment (class absent in training fold)
  keep <- !is.na(y_test)
  X_test <- X_test[keep, , drop = FALSE]
  y_test <- y_test[keep]   # IMPORTANT: no droplevels()

  set.seed(123 + k)
  mlp_cv <- nnet(
    x = as.matrix(X_train),
    y = class.ind(y_train),
    size = 10,
    decay = 1e-3,
    maxit = 300,
    softmax = TRUE,
    trace = FALSE,
    MaxNWts = 100000
  )

  prob <- predict(mlp_cv, as.matrix(X_test), type = "raw")
  pred <- colnames(prob)[max.col(prob)]
  pred <- factor(pred, levels = levels(y_train))

  cm <- confusionMatrix(pred, y_test)
  acc_vec[k]   <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)

```




# MLP with 20 PCs

```{r}
# sanity check
stopifnot(nrow(X100) == nrow(meta_treat))
stopifnot(length(y) == nrow(meta_treat))
table(y)
length(unique(meta_treat$sra_study))


```




```{r}
# split into train and test
set.seed(123)
studies <- unique(meta_treat$sra_study)
test_studies  <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_treat$sra_study %in% train_studies
test_idx  <- meta_treat$sra_study %in% test_studies

X_train <- X20[train_idx, , drop = FALSE]
X_test  <- X20[test_idx,  , drop = FALSE]

y_train <- droplevels(y[train_idx])
y_test  <- factor(y[test_idx], levels = levels(y_train))

dim(X_train); dim(X_test)
table(y_train); table(y_test)


```



```{r}
# Perform MLP

library(nnet)

set.seed(123)
mlp20 <- nnet(
  x = as.matrix(X_train),
  y = class.ind(y_train),
  size = 10,
  decay = 1e-3,
  maxit = 300,
  softmax = TRUE,
  trace = FALSE,
  MaxNWts = 100000
)

# --- 3) Predict + Confusion Matrix
prob_test <- predict(mlp20, as.matrix(X_test), type = "raw")
pred_test <- colnames(prob_test)[max.col(prob_test)]
pred_test <- factor(pred_test, levels = levels(y_train))   

cm_test <- confusionMatrix(pred_test, y_test)
cm_test

```



```{r}
# probabilities
prob_test <- predict(mlp20, as.matrix(X_test), type = "raw")
head(prob_test)

```



```{r}
# cross-validation MLP
library(caret)
library(nnet)

set.seed(123)

studies <- unique(meta_treat$sra_study)
folds <- createFolds(studies, k = 10, list = TRUE)

acc_vec   <- numeric(length(folds))
kappa_vec <- numeric(length(folds))

for (k in seq_along(folds)) {

  test_studies <- studies[folds[[k]]]
  test_idx  <- meta_treat$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X20[train_idx, , drop = FALSE]
  X_test  <- X20[test_idx,  , drop = FALSE]

  y_train <- droplevels(y[train_idx])
  y_test  <- factor(y[test_idx], levels = levels(y_train))  # align to train

  # remove test rows that became NA after alignment (class absent in training fold)
  keep <- !is.na(y_test)
  X_test <- X_test[keep, , drop = FALSE]
  y_test <- y_test[keep]   # IMPORTANT: no droplevels()

  set.seed(123 + k)
  mlp_cv <- nnet(
    x = as.matrix(X_train),
    y = class.ind(y_train),
    size = 10,
    decay = 1e-3,
    maxit = 300,
    softmax = TRUE,
    trace = FALSE,
    MaxNWts = 100000
  )

  prob <- predict(mlp_cv, as.matrix(X_test), type = "raw")
  pred <- colnames(prob)[max.col(prob)]
  pred <- factor(pred, levels = levels(y_train))

  cm <- confusionMatrix(pred, y_test)
  acc_vec[k]   <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)

```



# MLP with 50 PCs 

```{r}
# sanity check
stopifnot(nrow(X100) == nrow(meta_treat))
stopifnot(length(y) == nrow(meta_treat))
table(y)
length(unique(meta_treat$sra_study))


```




```{r}
# split into train and test
set.seed(123)
studies <- unique(meta_treat$sra_study)
test_studies  <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_treat$sra_study %in% train_studies
test_idx  <- meta_treat$sra_study %in% test_studies

X_train <- X50[train_idx, , drop = FALSE]
X_test  <- X50[test_idx,  , drop = FALSE]

y_train <- droplevels(y[train_idx])
y_test  <- factor(y[test_idx], levels = levels(y_train))

dim(X_train); dim(X_test)
table(y_train); table(y_test)

```





```{r}
# Perform MLP

library(nnet)

set.seed(123)
mlp50 <- nnet(
  x = as.matrix(X_train),
  y = class.ind(y_train),
  size = 10,
  decay = 1e-3,
  maxit = 300,
  softmax = TRUE,
  trace = FALSE,
  MaxNWts = 100000
)

# --- 3) Predict + Confusion Matrix
prob_test <- predict(mlp50, as.matrix(X_test), type = "raw")
pred_test <- colnames(prob_test)[max.col(prob_test)]
pred_test <- factor(pred_test, levels = levels(y_train))   

cm_test <- confusionMatrix(pred_test, y_test)
cm_test

```




```{r}
# probabilities
prob_test <- predict(mlp50, as.matrix(X_test), type = "raw")
head(prob_test)

```



```{r}
# cross-validation MLP
library(caret)
library(nnet)

set.seed(123)

studies <- unique(meta_treat$sra_study)
folds <- createFolds(studies, k = 10, list = TRUE)

acc_vec   <- numeric(length(folds))
kappa_vec <- numeric(length(folds))

for (k in seq_along(folds)) {

  test_studies <- studies[folds[[k]]]
  test_idx  <- meta_treat$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X50[train_idx, , drop = FALSE]
  X_test  <- X50[test_idx,  , drop = FALSE]

  y_train <- droplevels(y[train_idx])
  y_test  <- factor(y[test_idx], levels = levels(y_train))  # align to train

  # remove test rows that became NA after alignment (class absent in training fold)
  keep <- !is.na(y_test)
  X_test <- X_test[keep, , drop = FALSE]
  y_test <- y_test[keep]   # IMPORTANT: no droplevels()

  set.seed(123 + k)
  mlp_cv <- nnet(
    x = as.matrix(X_train),
    y = class.ind(y_train),
    size = 10,
    decay = 1e-3,
    maxit = 300,
    softmax = TRUE,
    trace = FALSE,
    MaxNWts = 100000
  )

  prob <- predict(mlp_cv, as.matrix(X_test), type = "raw")
  pred <- colnames(prob)[max.col(prob)]
  pred <- factor(pred, levels = levels(y_train))

  cm <- confusionMatrix(pred, y_test)
  acc_vec[k]   <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)


```



# MLP with 100 Pcs

```{r}
# sanity check
stopifnot(nrow(X100) == nrow(meta_treat))
stopifnot(length(y) == nrow(meta_treat))
table(y)
length(unique(meta_treat$sra_study))


```




```{r}
# split into train and test
set.seed(123)
studies <- unique(meta_treat$sra_study)
test_studies  <- sample(studies, size = round(0.2 * length(studies)))
train_studies <- setdiff(studies, test_studies)

train_idx <- meta_treat$sra_study %in% train_studies
test_idx  <- meta_treat$sra_study %in% test_studies

X_train <- X100[train_idx, , drop = FALSE]
X_test  <- X100[test_idx,  , drop = FALSE]

y_train <- droplevels(y[train_idx])
y_test  <- factor(y[test_idx], levels = levels(y_train))

dim(X_train); dim(X_test)
table(y_train); table(y_test)

```




```{r}
# Perform MLP

library(nnet)

set.seed(123)
mlp100 <- nnet(
  x = as.matrix(X_train),
  y = class.ind(y_train),
  size = 10,
  decay = 1e-3,
  maxit = 300,
  softmax = TRUE,
  trace = FALSE,
  MaxNWts = 100000
)

# --- 3) Predict + Confusion Matrix
prob_test <- predict(mlp100, as.matrix(X_test), type = "raw")
pred_test <- colnames(prob_test)[max.col(prob_test)]
pred_test <- factor(pred_test, levels = levels(y_train))   

cm_test <- confusionMatrix(pred_test, y_test)
cm_test


```




```{r}
# probabilities
prob_test <- predict(mlp100, as.matrix(X_test), type = "raw")
head(prob_test)

```


```{r}
# cross-validation MLP
library(caret)
library(nnet)

set.seed(123)

studies <- unique(meta_treat$sra_study)
folds <- createFolds(studies, k = 10, list = TRUE)

acc_vec   <- numeric(length(folds))
kappa_vec <- numeric(length(folds))

for (k in seq_along(folds)) {

  test_studies <- studies[folds[[k]]]
  test_idx  <- meta_treat$sra_study %in% test_studies
  train_idx <- !test_idx

  X_train <- X100[train_idx, , drop = FALSE]
  X_test  <- X100[test_idx,  , drop = FALSE]

  y_train <- droplevels(y[train_idx])
  y_test  <- factor(y[test_idx], levels = levels(y_train))  # align to train

  # remove test rows that became NA after alignment (class absent in training fold)
  keep <- !is.na(y_test)
  X_test <- X_test[keep, , drop = FALSE]
  y_test <- y_test[keep]   # IMPORTANT: no droplevels()

  set.seed(123 + k)
  mlp_cv <- nnet(
    x = as.matrix(X_train),
    y = class.ind(y_train),
    size = 10,
    decay = 1e-3,
    maxit = 300,
    softmax = TRUE,
    trace = FALSE,
    MaxNWts = 100000
  )

  prob <- predict(mlp_cv, as.matrix(X_test), type = "raw")
  pred <- colnames(prob)[max.col(prob)]
  pred <- factor(pred, levels = levels(y_train))

  cm <- confusionMatrix(pred, y_test)
  acc_vec[k]   <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)


```



# stupid way in MLP for 50 PCs

```{r}

library(nnet)
library(caret)

set.seed(123)

X50 <- as.data.frame(X50)
y   <- droplevels(y)

#  STUPID split MLP (sample-based 80/20) + Train/Test evaluation
set.seed(123)
train_idx <- createDataPartition(y, p = 0.8, list = FALSE)

X_train <- X50[train_idx, , drop = FALSE]
X_test  <- X50[-train_idx, , drop = FALSE]

y_train <- droplevels(y[train_idx])
y_test  <- factor(y[-train_idx], levels = levels(y_train))

set.seed(123)
mlp50_stupid <- nnet(
  x = as.matrix(X_train),
  y = class.ind(y_train),
  size = 10,
  decay = 1e-3,
  maxit = 300,
  softmax = TRUE,
  trace = FALSE,
  MaxNWts = 100000
)

prob_test <- predict(mlp50_stupid, as.matrix(X_test), type = "raw")
pred_test <- colnames(prob_test)[max.col(prob_test)]
pred_test <- factor(pred_test, levels = levels(y_train))

cm_test <- confusionMatrix(pred_test, y_test)
cm_test

```


```{r}
# STUPID 10-fold CV MLP (sample-based)

set.seed(123)
folds <- createFolds(y, k = 10, list = TRUE)

acc_vec   <- numeric(length(folds))
kappa_vec <- numeric(length(folds))

for (k in seq_along(folds)) {

  test_idx  <- folds[[k]]
  train_idx <- setdiff(seq_len(nrow(X50)), test_idx)

  X_train <- X50[train_idx, , drop = FALSE]
  X_test  <- X50[test_idx,  , drop = FALSE]

  y_train <- droplevels(y[train_idx])
  y_test  <- factor(y[test_idx], levels = levels(y_train))

  # (rare) if any NA after aligning levels, drop them
  keep <- !is.na(y_test)
  X_test <- X_test[keep, , drop = FALSE]
  y_test <- y_test[keep]

  set.seed(123 + k)
  mlp_cv <- nnet(
    x = as.matrix(X_train),
    y = class.ind(y_train),
    size = 10,
    decay = 1e-3,
    maxit = 300,
    softmax = TRUE,
    trace = FALSE,
    MaxNWts = 100000
  )

  prob <- predict(mlp_cv, as.matrix(X_test), type = "raw")
  pred <- colnames(prob)[max.col(prob)]
  pred <- factor(pred, levels = levels(y_train))

  cm <- confusionMatrix(pred, y_test)
  acc_vec[k]   <- as.numeric(cm$overall["Accuracy"])
  kappa_vec[k] <- as.numeric(cm$overall["Kappa"])
}

mean(acc_vec); sd(acc_vec)
mean(kappa_vec); sd(kappa_vec)


```


```{r}
library(ggplot2)

mlp_compare_df <- data.frame(
  Split = c("SMART (study-aware)", "STUPID (sample-based)"),
  Test_Accuracy = c(0.2576, 0.8990),
  CV_Mean_Accuracy = c(0.5115, 0.8808)
)

mlp_compare_df

```


```{r}
ggplot(mlp_compare_df, aes(x = Split)) +
  geom_point(aes(y = Test_Accuracy, shape = "Test accuracy"), size = 4) +
  geom_point(aes(y = CV_Mean_Accuracy, shape = "CV mean accuracy"), size = 4) +
  geom_line(aes(y = Test_Accuracy, group = 1, linetype = "Test accuracy")) +
  geom_line(aes(y = CV_Mean_Accuracy, group = 1, linetype = "CV mean accuracy")) +
  labs(
    title = "MLP (50 PCs): SMART vs STUPID split",
    y = "Accuracy",
    x = "Split strategy",
    shape = "",
    linetype = ""
  ) +
  theme_minimal()

```


```{r}
library(ggplot2)

mlp_comp_df <- data.frame(
  split = factor(c("SMART (study-aware)", "STUPID (sample-based)"),
                 levels = c("SMART (study-aware)", "STUPID (sample-based)")),
  cv_mean = c(0.511451, 0.8807659),
  cv_sd   = c(0.1993211, 0.02216319),
  test    = c(0.2576,    0.8990)
)

ggplot(mlp_comp_df, aes(x = split)) +
  geom_point(aes(y = cv_mean, shape = "CV mean"), size = 3) +
  geom_errorbar(aes(ymin = cv_mean - cv_sd, ymax = cv_mean + cv_sd),
                width = 0.15) +
  geom_point(aes(y = test, shape = "Test"), size = 3) +
  scale_shape_manual(values = c("CV mean" = 16, "Test" = 17)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "MLP (50 PCs): SMART vs STUPID (Accuracy)",
    x = "Split strategy",
    y = "Accuracy",
    shape = ""
  ) +
  theme_minimal()

```


```{r}
```


```{r}
```


```{r}
```



```{r}
#comparing RF and MLP 50 PCs
comparison_50PC <- data.frame(
  Model = c("Random Forest (50 PCs)", "MLP (50 PCs)"),
  CV_Accuracy = c(0.428, 0.511),
  CV_Kappa = c(0.188, 0.173)
)

comparison_50PC

```




```{r}
# comparing RF and MLP over PCs
comparison_all <- data.frame(
  PCs = rep(c(10,20,50,100), 2),
  Model = rep(c("Random Forest", "MLP"), each = 4),
  CV_Accuracy = c( 0.456, 0.461, 0.428, 0.444,   # RF
    0.313, 0.409, 0.511, 0.417    # MLP
  ),
  CV_Kappa = c(
    0.105, 0.104, 0.075, 0.072,   # RF
    -0.0098, -0.0102, 0.1727, 0.1201  # MLP
  )
)

comparison_all

```



```{r}
library(ggplot2)

ggplot(comparison_all, aes(x = PCs, y = CV_Accuracy, color = Model)) +
  geom_point(size = 3) +
  geom_line(size = 1) +
  theme_minimal() +
  labs(
    title = "RF vs MLP Performance across PCs (Treatment)",
    y = "Cross-Validated Accuracy"
  )

```


# creating error bars

```{r}
#creating data frame with values
results_df <- data.frame(
  PCs = c(10, 20, 50, 100),

  RF_acc_mean = c(0.4568, 0.4615, 0.4284, 0.4438),
  RF_acc_sd   = c(0.1361, 0.1488, 0.1446, 0.1646),
  RF_kap_mean = c(0.1052, 0.1869, 0.2337, 0.1198),
  RF_kap_sd   = c(0.1440, 0.2262, 0.0809, 0.1198),

  MLP_acc_mean = c(0.3129, 0.4091, 0.5115, 0.4176),
  MLP_acc_sd   = c(0.1440, 0.1750, 0.1993, 0.1997),
  MLP_kap_mean = c(-0.0098, -0.0102, 0.1727, 0.1201),
  MLP_kap_sd   = c(0.1567, 0.2543, 0.2202, 0.1621)
)

```





```{r}
library(tidyr)
library(dplyr)
#converting to long format
acc_df <- results_df %>%
  select(PCs, RF_acc_mean, RF_acc_sd, MLP_acc_mean, MLP_acc_sd) %>%
  pivot_longer(-PCs, names_to = c("Model","Metric","Stat"), names_sep = "_") %>%
  pivot_wider(names_from = Stat, values_from = value)

kap_df <- results_df %>%
  select(PCs, RF_kap_mean, RF_kap_sd, MLP_kap_mean, MLP_kap_sd) %>%
  pivot_longer(-PCs, names_to = c("Model","Metric","Stat"), names_sep = "_") %>%
  pivot_wider(names_from = Stat, values_from = value)

```




```{r}
library(ggplot2)

# Accuracy Plot (RF and MLP over PCs)
ggplot(acc_df, aes(x = PCs, y = mean, group = Model, shape = Model, linetype = Model)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 3) +
  ylim(0, 1) +
  labs(title = "RF vs MLP (Treatment): CV Accuracy over PCs",
       x = "Number of PCs",
       y = "CV Accuracy",
       shape = "Model",
       linetype = "Model") +
  theme_minimal()

```


```{r}
# Kappa plot ( RF and MLP over PCs)

ggplot(kap_df, aes(x = PCs, y = mean, group = Model, shape = Model, linetype = Model)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 3) +
  labs(title = "RF vs MLP (Treatment): CV Kappa over PCs",
       x = "Number of PCs",
       y = "CV Kappa",
       shape = "Model",
       linetype = "Model") +
  theme_minimal()


```


**FINAL MODEL TRAINING and DEPLOYMENT**

# FINAL MODEL TRAINING and DEPLYMENT with 100 PCs on unknown treatments to predict their classes (control, abiotic, biotic)



```{r}
# till now i have performed rf and cross validation for known treatments with 10,20,50 and 100 PCs. now i want to perform random forest and cross validation for on unknown treatments by using 100 PCs 
# now am doing final training on all known treatment samples with 100 PCs and then will predict on unknown treatment samples.

## FINAL STEP (Treatment):
## Train RF on ALL known samples (100 PCs) and predict on unknown samples

library(randomForest)
library(dplyr)

# Make sure meta2 has sra_study + treatment (character)
meta_all <- meta2 %>%
  mutate(
    sra_study = sra_study,                 # keep as is (or rename if needed)
    treatment = as.character(treatment)
  )

```




```{r}
# 1) Create treatment_group (THIS is the missing part that caused your error)
meta_all <- meta_all %>%
  mutate(
    treatment_group = case_when(
      treatment == "control" ~ "control",
      treatment %in% c("abiotic stress","drought","salt","heat","cold","oxidative stress") ~ "abiotic",
      treatment %in% c("pathogen","symbiotic") ~ "biotic",
      TRUE ~ NA_character_   # everything else = unknown for this final model
    )
  )

```




```{r}
# 2) Build PC matrix for ALL samples (must match meta row order used in PCA)
X100_all <- as.data.frame(pca_res$x[, 1:100, drop = FALSE])
stopifnot(nrow(X100_all) == nrow(meta_all))

```




```{r}
# 3) Define known vs unknown (unknown includes NA in treatment_group)
known_idx   <- !is.na(meta_all$treatment_group)
unknown_idx <-  is.na(meta_all$treatment_group)

```




```{r}
# sanity checks (important)
table(meta_all$treatment_group, useNA = "ifany")
sum(known_idx); sum(unknown_idx)

```


# here 2088 are known samples with treatment_group labels (control, abiotic, biotic) and 8008 are unknown samples with NA in treatment_group that we want to predict 
# total samples = 10096 (2088 known + 8008 unknown) which are total my metadata samples



```{r}
# 4) Train FINAL RF on ALL known samples (3-class)
X_train_all <- X100_all[known_idx, , drop = FALSE]
y_train_all <- droplevels(factor(meta_all$treatment_group[known_idx]))

table(y_train_all)   # should show abiotic/biotic/control (>=2 classes!)

rf_final <- randomForest(
  x = X_train_all,
  y = y_train_all,
  ntree = 500,
  classwt = 1 / table(y_train_all)
)

```



```{r}
# 5) Predict unknown samples (labels + probabilities)
X_unknown <- X100_all[unknown_idx, , drop = FALSE]

pred_unknown <- predict(rf_final, X_unknown, type = "response")
prob_unknown <- predict(rf_final, X_unknown, type = "prob")

```




# lets explore the predictions for the unknown samples to see what the model is predicting .

```{r}
# how many unknown you predicted?
length(pred_unknown)
nrow(prob_unknown)

```



```{r}
# predicted class distribution (unknown set)
# NOTE: this is just the distribution of predicted classes for the unknown samples, not a performance metric since we don't have true labels for these
# this shows how much percentage of unknown samples were predicted as each class (control, abiotic, biotic)

prop.table(table(pred_unknown))

```



```{r}
# quick look at probabilities
head(prob_unknown)

```


```{r}
# confidence summary (max prob per sample)
# this gives an idea of how confident the model is in its predictions for the unknown samples (higher max prob = more confident)
# in simple terms, this shows the distribution of the highest predicted probability across all unknown samples, which can indicate how strongly the model is leaning towards its predicted class for those samples 

conf <- apply(prob_unknown, 1, max)
summary(conf)
quantile(conf, c(0.1, 0.25, 0.5, 0.75, 0.9))

```



```{r}
# show the 20 most confident predictions (with sample IDs)
# this lists the top 20 unknown samples that the model is most confident about (highest max probability), along with their predicted class and confidence score confidence score 1 is very high confidence, while values closer to 0.5 indicate more uncertainty in the prediction (since it's a 3-class problem, the baseline random guess would be around 0.33) 
# for example, if a sample has a confidence score of 0.9 for being classified as "abiotic", it means the model is very confident that this sample belongs to the "abiotic" class, while a score of 0.4 would indicate much less confidence in that prediction.

top <- order(conf, decreasing = TRUE)[1:20]
data.frame(
  sample = rownames(prob_unknown)[top],
  pred   = pred_unknown[top],
  conf   = conf[top]
)

```




```{r}
# show the 20 least confident predictions
bot <- order(conf, decreasing = FALSE)[1:20]
data.frame(
  sample = rownames(prob_unknown)[bot],
  pred   = pred_unknown[bot],
  conf   = conf[bot]
)

```



```{r}
# check if model is just predicting mostly 'control'
table(pred_unknown)

```



```{r}
# here we can also check the agreement of predictions across multiple runs to see if the model is stable in its predictions for the unknown samples (since we don't have true labels, this is a way to assess consistency)  
# here we train the same RF model multiple times with different random seeds and see if it gives the same predictions for the unknown samples each time. if the agreement is high (close to 1), it means the model is consistently predicting the same class for those unknown samples, which can increase our confidence in those predictions. if the agreement is low, it suggests that the model's predictions for the unknown samples are more variable and less reliable.

#different random seeds means different random subsets of trees in the RF, which can lead to slightly different predictions. if the model is very confident and stable in its predictions for the unknown samples, we would expect high agreement across runs. if the model is uncertain about those samples, we might see more variability in the predicted classes across runs, leading to lower agreement.
 
# one line about this step: "To further assess the stability of our predictions for the unknown samples, we can train the random forest model multiple times with different random seeds and check how consistent the predicted classes are across these runs. High agreement would suggest that the model is confidently classifying those unknown samples, while low agreement might indicate uncertainty in those predictions."

pred_list <- list()

for (s in 1:5) {
  set.seed(100 + s)
  
  rf_tmp <- randomForest(
    x = X_train_all,
    y = y_train_all,
    ntree = 500,
    classwt = 1 / table(y_train_all)
  )
  
  pred_list[[s]] <- predict(rf_tmp, X_unknown)
}

# agreement rate
agree_matrix <- do.call(cbind, pred_list)
mean(apply(agree_matrix, 1, function(x) length(unique(x)) == 1))

```
#98.25% of unknown samples received the SAME predicted class across 5 different RF runs





#final modeland deployment for unknown samples with 50 PCs
#(since 100 PCs had lower CV performance than 50 PCs, we can try the same final prediction approach with 50 PCs to see if it gives more confident predictions for the unknown samples)


```{r}
# 2) Build PC matrix for ALL samples (must match meta row order used in PCA)
X50_all <- as.data.frame(pca_res$x[, 1:50, drop = FALSE])
stopifnot(nrow(X50_all) == nrow(meta_all))

```




```{r}
# 3) Define known vs unknown (unknown includes NA in treatment_group)
known_idx   <- !is.na(meta_all$treatment_group)
unknown_idx <-  is.na(meta_all$treatment_group)

```





```{r}
# 4) Train FINAL RF on ALL known samples (3-class)
X_train_all <- X50_all[known_idx, , drop = FALSE]
y_train_all <- droplevels(factor(meta_all$treatment_group[known_idx]))

table(y_train_all)   # should show abiotic/biotic/control (>=2 classes!)

rf_final <- randomForest(
  x = X_train_all,
  y = y_train_all,
  ntree = 500,
  classwt = 1 / table(y_train_all)
)
```


```{r}
# 5) Predict unknown samples (labels + probabilities)
X_unknown <- X50_all[unknown_idx, , drop = FALSE]

pred_unknown <- predict(rf_final, X_unknown, type = "response")
prob_unknown <- predict(rf_final, X_unknown, type = "prob")


```




```{r}
# predicted class distribution (unknown set)
# this is just the distribution of predicted classes for the unknown samples, not a performance metric since we don't have true labels for these 
# this shows how much percentage of unknown samples were predicted as each class (control, abiotic, biotic)
prop.table(table(pred_unknown))

```



```{r}
#check if model is just predicting mostly 'control'
table(pred_unknown)

```


```{r}
#confidence summary (max prob per sample)
conf <- apply(prob_unknown, 1, max)
summary(conf)
quantile(conf, c(0.1, 0.25, 0.5, 0.75, 0.9))  

```



```{r}
# check the agreement of predictions across multiple runs to see if the model is stable in its predictions for the unknown samples (since we don't have true labels, this is a way to assess consistency)  
pred_list <- list() 
for (s in 1:5) {
  set.seed(100 + s)
  
  rf_tmp <- randomForest(
    x = X_train_all,
    y = y_train_all,
    ntree = 500,
    classwt = 1 / table(y_train_all)
  )
  
  pred_list[[s]] <- predict(rf_tmp, X_unknown)
}
# agreement rate
agree_matrix <- do.call(cbind, pred_list)
mean(apply(agree_matrix, 1, function(x) length(unique(x)) == 1))  

```




```{r}
# create empty columns first before assigning predictions (to avoid issues with NA vs factor levels)
meta_all$treatment_group_pred <- NA_character_
meta_all$treatment_group_conf <- NA_real_

# now assign predictions to unknown samples
meta_all$treatment_group_pred[unknown_idx] <- as.character(pred_unknown)
meta_all$treatment_group_conf[unknown_idx] <- conf

# # save full updated metadata
# write.csv(meta_all,
#           "metadata_with_treatment_predictions.csv",
#           row.names = FALSE)

```





```{r}
# check the columns in the updated metadata
colnames(meta_all)

```




```{r}  
# quick look at the updated metadata with predictions
head(meta_all)

```




```{r}  
# check the distribution of predicted classes in the metadata (should match what we saw in prop.table earlier)
table(meta_all$treatment_group_pred, useNA = "ifany")
summary(meta_all$treatment_group_conf)

```


```{r}  
```


```{r}  
